---
title: "BSS effort-bias term (`b`): predictive approach"
subtitle: "DRAFT — proposed next steps, not reviewed"
date: last-modified
execute:
  warning: false
  message: false
  echo: false
format:
  html:
    page-layout: full
    code-tools: true
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    title-block-banner: "#8a6d1f"
---

```{r setup}
library(tidyverse)
library(here)
library(gt)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")
FIG_DIR <- file.path(OUT_DIR, "figures")

# Same presentation-layer pattern as the main report: reads outputs/ and
# recomputes nothing, so it cannot disagree with the pipeline and re-renders
# in seconds. See ../bss_bias_report.qmd.

read_out <- function(name) {
  p <- file.path(OUT_DIR, name)
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
}

fig_path <- function(name) {
  p <- file.path(FIG_DIR, paste0(name, ".png"))
  if (file.exists(p)) p else NULL
}

missing_note <- function(script) {
  cat(paste0(
    "::: {.callout-note}\n## Not yet generated\n",
    "Run `", script, "` and re-render. This section fills itself in.\n:::\n"
  ))
}

T2 <- read_out("bss_b_T2_variability.csv")
toy_recovery <- read_out("bss_b_toy_recovery.csv")
```

::: {.callout-important}
## Status: draft
This is **not** part of the reviewed working-session document. It holds the
material that goes beyond the first-order variability question — the empirical
predictive distributions and a worked demonstration that the method recovers a
known answer.

It is here so the group has something concrete to react to when deciding
whether to take the predictive approach further. The conceptual model it
implements is in §4 of the [main report](../bss_bias_report.html).
:::

## Why this exists

The main report answers *how variable is `b`*. This one takes the next step:
**if we cannot measure `b` in a given fishery-year, what should we use, and how
uncertain is it?**

Nothing here is settled. The point is to make the proposal concrete enough to
argue with.

## What the history implies for an unmeasured year

```{r}
#| results: asis
p <- fig_path("fig8_predictive_distribution")
if (is.null(p)) missing_note("analysis/bss_bias/06_variability_analysis.R") else cat(paste0("![](", p, ")\n"))
p2 <- fig_path("fig8b_predictive_distribution_trailer")
if (!is.null(p2)) cat(paste0("\n![](", p2, ")\n"))
```

The shaded band is the **95% prediction interval for a year we did not
observe**, built from that fishery's own history. It is the direct answer to
"what do we use when there is no census data, and how uncertain is it?"

- **Tight band** → historical `b` transfers; use the pooled estimate with its PI.
- **Wide band** → it does not transfer, and a single historical value would be
  false precision.

```{r}
if (!is.null(T2)) {
  T2 |>
    filter(!is.na(pi_lb)) |>
    transmute(
      Basin = basin, Fishery = fishery_type, Likelihood = bias_type,
      `Yrs used` = n_years_used,
      b = round(pooled_b, 3),
      `PI lo` = round(pi_lb, 3), `PI hi` = round(pi_ub, 3),
      `PI / CI width` = round(pi_width_ratio, 2),
      tau = round(tau, 3)
    ) |>
    gt(groupname_col = "Likelihood") |>
    tab_header(
      title = "Predictive distribution for an unobserved year",
      subtitle = "The interval a new year's b is expected to fall in, per fishery series."
    ) |>
    tab_footnote(
      footnote = "PI / CI width: how much wider the predictive interval is than the confidence interval on the series mean. It is the price of predicting a new year rather than describing past ones.",
      locations = cells_column_labels(columns = `PI / CI width`)
    ) |>
    opt_row_striping()
}
```

## Does the approach actually work?

Before trusting this on real numbers, here it is on a **simulated world where
the truth is known** — three rivers, seven years, a known pink effect and known
between-river and between-year variances.

```{r}
#| results: asis
for (nm in c("figA_toy_simulated_world", "figC_toy_variance_components", "figB_toy_prediction_check")) {
  p <- fig_path(nm)
  if (!is.null(p)) cat(paste0("![](", p, ")\n\n"))
}
if (is.null(fig_path("figA_toy_simulated_world"))) missing_note("analysis/bss_bias/05_predictive_toy.R")
```

```{r}
if (!is.null(toy_recovery)) {
  toy_recovery |>
    gt() |>
    tab_header(title = "Parameter recovery on simulated data",
               subtitle = "Known truth vs. what the method recovered") |>
    opt_row_striping()
}
```

The held-out check is the use case in miniature: drop the most recent year,
refit on the rest, and ask whether the predictive distribution covers the value
that was hidden. **If it does not work here, where the truth is known and the
model is correctly specified, it will not work on the real data.**

## Open questions for the group

1. **Which rung?** The ladder in §4 of the main report runs from "use last
   year's `b`" to a hierarchical re-specification inside the Stan model.
   Rungs 1–3 are a two-stage approximation; rung 4 is a model rewrite.
2. **Does the pink covariate earn its place?** It is a plausible mechanism with
   a thin design behind it — few odd years against few even ones, and only the
   fall fisheries can carry the test at all.
3. **Pool across rivers, or keep series independent?** Pooling lets a
   data-poor fishery borrow strength, at the cost of assuming the rivers are
   exchangeable in a way nobody has checked.
4. **What is good enough?** A prediction interval honestly reflecting the
   uncertainty may be wide. The question is whether a wide honest interval is
   more useful for management than a narrow assumed one.

## Reproducing this

```bash
Rscript analysis/bss_bias/05_predictive_toy.R
Rscript analysis/bss_bias/06_variability_analysis.R
quarto render analysis/bss_bias/report/draft/bss_bias_next_steps.qmd
```
