---
title: "Effort-index bias (`b`): what it does to estimated catch"
date: today
params:
  focus: "Stillaguamish"
format:
  html:
    embed-resources: true
    toc: false
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# ==============================================================================
# ONE DOCUMENT THAT GROWS WITH THE OUTPUTS.
#
# Everything down to "How close a borrowed b gets" rests on the b draws and the
# meta-analysis alone -- T7's pct_change_catch and the T6 calibration. Neither
# reads a season total, so that material renders the moment 07 has run and is
# sendable while a model sweep is still going.
#
# The effort-and-catch section at the end needs the production fits. It renders
# only when bss_b_T8_gear_split.csv exists AND T7 carries season totals, and is
# replaced by a short "what is coming" note otherwise. Gating on the data rather
# than maintaining a second document means the two halves cannot drift apart,
# and nothing above can be affected by a baseline that is missing or stale.
# ==============================================================================
library(tidyverse)
library(gt)
library(here)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")
source(here::here("analysis", "bss_bias", "common.R"))

need <- function(f) {
  p <- file.path(OUT_DIR, f)
  if (!file.exists(p)) stop(f, " not found -- run 07_catch_sensitivity.R first.", call. = FALSE)
  readr::read_csv(p, show_col_types = FALSE)
}

T7  <- need("bss_b_T7_direct_sensitivity.csv")
cal <- need("bss_b_T6_calibration.csv")

# Which series this brief is about. A PARAM, not a constant, so one document
# renders a per-basin brief for each collaborator group:
#
#   quarto render 11_share_brief.qmd -P focus:Stillaguamish ...
#   quarto render 11_share_brief.qmd -P focus:Snohomish ...
#
# NOTE what is NOT subset by it: the backtest below. How well a borrowed b
# predicts is a question about the METHOD, and any one basin contributes only a
# handful of the 43 held-out tests. Scoring the method on those alone would be a
# weaker claim resting on a smaller sample, so the reliability figures stay
# across all three basins and the focal basin's own rows are called out within
# them.
FOCUS_RE <- params$focus

# One row per fishery-year x b term: b as fitted, the ends of that series' own
# plausible range, and the percent change in catch at each end.
#
# pct_change_catch, NOT pct_change_catch_total: the first is 100*(b_fitted/b_alt - 1),
# which needs no season total. The second divides by C_sum and is NA until the
# fits land, so it belongs to the in-fish section at the end -- this table has
# to render either way.
#
# Pivot on a short stable key rather than the display label, so rewording a
# label for the reader does not move a column name.
wide <- T7 |>
  filter(tier_kind %in% c("anchor", "empirical")) |>
  mutate(key = case_when(
    tier_kind == "anchor"        ~ "fitted",
    str_detect(tier, "^low end") ~ "low",
    str_detect(tier, "^high end")~ "high",
    TRUE                         ~ NA_character_
  )) |>
  filter(!is.na(key)) |>
  distinct(fishery_type, fishery_name, year_start, bias_type, key,
           b_alt, pct_change_catch) |>
  pivot_wider(names_from = key, values_from = c(b_alt, pct_change_catch)) |>
  transmute(
    fishery_type, fishery_name, year_start, bias_type,
    b_fitted = b_alt_fitted,
    b_low    = b_alt_low,
    b_high   = b_alt_high,
    pct_low  = pct_change_catch_low,     # b at the LOW end -> catch UP
    pct_high = pct_change_catch_high     # b at the HIGH end -> catch DOWN
  ) |>
  filter(!is.na(b_low), !is.na(b_high)) |>
  filter(str_detect(fishery_type, FOCUS_RE)) |>
  arrange(fishery_type, year_start, bias_type)

if (nrow(wide) == 0) stop("No fishery_type matches focus = '", FOCUS_RE, "'.", call. = FALSE)

# The series name for prose and headings, taken from the data rather than passed
# in. Guarantees the label matches what the filter actually kept, and keeps the
# only command-line param a single word with no quoting hazard.
SERIES_LABEL <- {
  ft <- sort(unique(wide$fishery_type))
  if (length(ft) == 1) ft else paste(ft, collapse = " / ")
}

# ------------------------------------------------------------------------------
# Season totals, if the production fits have landed
# ------------------------------------------------------------------------------

# est_cg is the full species_lifestage_finmark_fate string. Collaborators do not
# need to read a regex, so it is labelled by target species.
group_label <- function(x) {
  dplyr::case_when(
    stringr::str_starts(x, "Chinook") ~ "Chinook encounters",
    stringr::str_starts(x, "Coho")    ~ "Coho harvest",
    is.na(x)                          ~ NA_character_,
    TRUE                              ~ stringr::str_extract(x, "^[^_]+")
  )
}

t8_path <- file.path(OUT_DIR, "bss_b_T8_gear_split.csv")
T8 <- if (file.exists(t8_path)) readr::read_csv(t8_path, show_col_types = FALSE) else NULL

# At the anchor tier b_alt == b_fitted, so both gear factors are 1 and the row
# is the fit's own season total. Either bias_type gives the same numbers there.
as_fitted <- if (is.null(T8)) NULL else T8 |>
  filter(tier_kind == "anchor", bias_type == "vehicle",
         str_detect(fishery_type, FOCUS_RE)) |>
  mutate(group = group_label(est_cg),
         boat_share_catch = catch_boat / catch_total) |>
  arrange(fishery_name, group)

# The same sensitivity as the table above, but in fish. pct_change_catch_total
# here, NOT pct_change_catch: the first is measured against the whole season
# total and is what a reader quotes; they differ for trailer, where only the
# boat component moves.
fish <- T7 |>
  filter(tier_kind %in% c("anchor", "empirical"),
         str_detect(fishery_type, FOCUS_RE)) |>
  mutate(group = group_label(est_cg),
         key = case_when(
           tier_kind == "anchor"         ~ "fitted",
           str_detect(tier, "^low end")  ~ "low",
           str_detect(tier, "^high end") ~ "high",
           TRUE                          ~ NA_character_
         )) |>
  filter(!is.na(key)) |>
  select(fishery_type, fishery_name, year_start, bias_type, group, key,
         catch_estimate, pct_change_catch_total) |>
  pivot_wider(names_from = key,
              values_from = c(catch_estimate, pct_change_catch_total)) |>
  transmute(
    fishery_type, fishery_name, year_start, bias_type, group,
    fitted    = catch_estimate_fitted,
    if_b_low  = catch_estimate_low,
    if_b_high = catch_estimate_high,
    pct_low   = pct_change_catch_total_low,
    pct_high  = pct_change_catch_total_high
  ) |>
  filter(!is.na(fitted)) |>
  arrange(fishery_type, year_start, group, bias_type)

# Both conditions, not either: the gear split without season totals gives a
# table of NAs, and season totals without the gear split would report the
# trailer term as if it moved the whole fishery -- which is what produced the
# -58% / +260% rows before the gear totals existed.
HAVE_FISH <- !is.null(as_fitted) && nrow(as_fitted) > 0 && nrow(fish) > 0

```

`b` is the correction between what an index count sees and how many anglers are
actually out there. Measuring it needs census (tie-in) counts. Our key question
is whether a `b` carried over from previous years can stand in for a
measured one in a year where census counts are not collected, and if it can,
how much the catch estimate moves due to changes in `b`.

::: {.callout-note collapse="true"}
## Why catch is proportional to 1/`b`

Three observations enter the effort side of the model, and only one carries
$b$ (notation as in the full report):

$$
\begin{aligned}
E_s &\sim \text{Poisson}\big(\lambda\big)
  &&\text{census (tie-in): everything in the block}\\[2pt]
V_I &\sim \text{Poisson}\big(\lambda \cdot R_V \cdot b_1\big)
  &&\text{vehicle index: the counted sites}\\[2pt]
V_A &\sim \text{Binomial}\big(A_A,\; R_V\big)
  &&\text{interviews: } A_A \text{ anglers, } V_A \text{ vehicles}
\end{aligned}
$$

**With no census**, the first line is absent and $\lambda$ -- anglers present --
is pinned only by the index counts:

$$
\lambda \;=\; \frac{V_I}{R_V \;\cdot\; b_1}
$$

Catch is effort $\times$ trip length $\times$ catch rate, and the catch rate
$c$ comes from the interviews alone -- it never sees $b$:

$$
C \;=\; \lambda \cdot L \cdot c
\;=\; \frac{V_I \cdot L \cdot c}{R_V \cdot b_1}
$$

$V_I$ is the count itself, $L$ is day length, and $R_V$ and $c$ both come from
the interviews. Every term but $b_1$ is fixed by something other than $b$, so

$$
C \;\propto\; \frac{1}{b_1}
$$

Effort is $E = \lambda \cdot L$, which carries the same $1/b_1$, so the catch
rate $C/E = c$ is **unchanged** by $b$. That is why `b` rescales the fishery
without changing the rate at which fish are caught.

:::

## What the plausible range of `b` does to estimated catch

Low and high below are the ends of the **95% prediction interval** for `b` in
that series, based on an earlier part of this analysis.
This can be interpreted as the range a new, unmeasured year could be expected to fall in, derived from
other years of that same fishery. Essentially, they are what `b` could be given what has been observed
in the past.

```{r table}
# The two index channels do not observe the same anglers. The vehicle count sees
# every angler's vehicle, so b[1] rescales the whole fishery. Trailers per BANK
# angler go to ~0 in the interviews, so the trailer count sees boat anglers
# essentially alone and b[2] moves the boat-caught portion. Saying so in the
# row label keeps the percent honest without needing the gear totals, which is
# exactly what part 2 supplies.
wide |>
  mutate(
    term = if_else(bias_type == "vehicle",
                   "vehicle -- whole fishery",
                   "trailer -- boat-caught portion"),
    b_range = sprintf("%.2f to %.2f", b_low, b_high),
    yr = as.character(year_start)
  ) |>
  select(fishery_type, yr, term, b_fitted, b_range, pct_low, pct_high) |>
  gt(groupname_col = "fishery_type") |>
  cols_label(yr = "Year", term = "b term", b_fitted = "b as fitted",
             b_range = "Plausible range for b",
             pct_low = "If b is at the low end",
             pct_high = "If b is at the high end") |>
  fmt_number(b_fitted, decimals = 2) |>
  fmt_number(c(pct_low, pct_high), decimals = 0, pattern = "{x}%", force_sign = TRUE) |>
  cols_align("right", c(b_fitted, b_range, pct_low, pct_high)) |>
  sub_missing(everything(), missing_text = "--") |>
  tab_options(table.font.size = 13, data_row.padding = px(4),
              row_group.font.weight = "bold") |>
  tab_source_note(
    paste0("Change in estimated catch. The percentage is the same for every species and ",
           "catch group, because b moves effort and leaves the catch rate alone. ",
           "Vehicle rows move the whole fishery; trailer rows move the boat-caught ",
           "portion, so their effect on the season total differs by however much of ",
           "the fishery is fished from the bank.")
  )
```

```{r figure, fig.width = 9, fig.height = 6}
wide |>
  mutate(term = if_else(bias_type == "vehicle",
                        "vehicle (whole fishery)", "trailer (boat-caught portion)")) |>
  ggplot(aes(y = fct_rev(factor(fishery_name)), color = term)) +
  geom_vline(xintercept = 0, color = INK_SECOND, linewidth = 0.5, linetype = "22") +
  geom_linerange(aes(xmin = pct_high, xmax = pct_low),
                 position = position_dodge(width = 0.55), linewidth = 2.2, alpha = 0.9) +
  scale_x_continuous(labels = function(x) paste0(ifelse(x > 0, "+", ""), x, "%")) +
  scale_color_manual(values = c("vehicle (whole fishery)" = CAT[["blue"]],
                                "trailer (boat-caught portion)" = CAT[["orange"]]),
                     name = NULL) +
  labs(title = "Change in estimated catch across the plausible range of b",
       x = "Change in estimated catch", y = NULL) +
  theme_bss()
```

Each bar spans what the catch estimate would be at the two ends of the plausible
range for `b`. The right-hand end is the high `b` (effort, and thus catch, revised down), the
left-hand end the low `b` (effort, and thus, catch revised up).

## How close a borrowed `b` gets

```{r calibration-line, results = "asis"}
# Deliberately NOT subset to FOCUS_RE -- see the note in the setup chunk. How
# well the method predicts is scored on every series, then the focal basin's own
# rows are reported inside that.
v <- cal |> filter(bias_type == "vehicle")
t <- cal |> filter(bias_type == "trailer")
cat(sprintf(
  paste0("Each fishery-year with a measured `b` was predicted from the **other** years of ",
         "its own series and nothing else, then compared against what was actually ",
         "measured -- %d such tests across the three basins. A predicted **vehicle** `b` ",
         "landed within **%.0f%%** of the measured value in the typical year, worst case ",
         "%.0f%%. The **trailer** term is less reliable: typically ",
         "**%.0f%%** off, worst case **%.0f%%**.\n\n"),
  sum(cal$n_years_tested),
  median(v$median_abs_pct_err), max(abs(v$worst_pct_err)),
  median(t$median_abs_pct_err), max(abs(t$worst_pct_err))
))

foc <- cal |> filter(str_detect(fishery_type, FOCUS_RE))
fv  <- foc |> filter(bias_type == "vehicle")
ft  <- foc |> filter(bias_type == "trailer")
# A basin missing one of the two terms would otherwise print NA and -Inf into a
# document going to co-managers. Say nothing rather than that.
if (nrow(fv) > 0 && nrow(ft) > 0) cat(sprintf(
  paste0("%s contributes %d of those tests. Its **vehicle** term came in ",
         "inside the interval %d of %d times, typically **%.0f%%** off the measured ",
         "value (worst %.0f%%); its **trailer** term %d of %d, typically **%.0f%%** off ",
         "(worst %.0f%%).\n"),
  SERIES_LABEL, sum(foc$n_years_tested),
  sum(fv$n_inside_95_pi), sum(fv$n_years_tested),
  median(fv$median_abs_pct_err), max(abs(fv$worst_pct_err)),
  sum(ft$n_inside_95_pi), sum(ft$n_years_tested),
  median(ft$median_abs_pct_err), max(abs(ft$worst_pct_err))
))
```

```{r follow-up, eval = !HAVE_FISH, results = "asis"}
# Their words, unchanged, for the version that goes out before the fits land.
cat(r"---(
## Follow up

Season total catch estimates by bank and boat gear strata, are being fitted
overnight. These updated results will put the effect in
fish, and will also help illustrate the magnitude of bank/boat angling effort and contextualize
their respective b terms.
)---")
```

```{r gear-intro, eval = HAVE_FISH, results = "asis"}
# Headings and prose have to be conditional too, so they go through cat() in an
# asis chunk. r"---( )---" is a raw string: no backslash escaping, which matters
# for the LaTeX below.
cat(r"---(
## Estimated effort and catch by gear

The percentages above hold for every species and catch group. Below they are
put in fish, using the fitted season totals, and split between bank and boat.
)---")
```

```{r fitted-totals, eval = HAVE_FISH}
as_fitted |>
  select(fishery_name, group, catch_bank, catch_boat, catch_total,
         effort_bank, effort_boat, boat_share_catch) |>
  gt(groupname_col = "fishery_name") |>
  cols_label(group = "Catch group",
             catch_bank = "Bank", catch_boat = "Boat", catch_total = "Total",
             effort_bank = "Bank", effort_boat = "Boat",
             boat_share_catch = "Boat share") |>
  tab_spanner("Catch (fish)", c(catch_bank, catch_boat, catch_total)) |>
  tab_spanner("Effort (angler-hours)", c(effort_bank, effort_boat)) |>
  fmt_number(c(catch_bank, catch_boat, catch_total), decimals = 0) |>
  fmt_number(c(effort_bank, effort_boat), decimals = 0) |>
  fmt_percent(boat_share_catch, decimals = 0) |>
  sub_missing(everything(), missing_text = "--") |>
  cols_align("right", c(catch_bank, catch_boat, catch_total,
                        effort_bank, effort_boat, boat_share_catch)) |>
  tab_options(table.font.size = 13, data_row.padding = px(4),
              row_group.font.weight = "bold") |>
  tab_source_note(
    paste0("Posterior medians. Gear totals are the model's own -- the daily, ",
           "section-level draws summed within gear, not a share applied afterwards.")
  )
```

```{r gear-math, eval = HAVE_FISH, results = "asis"}
cat(r"---(
::: {.callout-note collapse="true"}
## How the two terms divide the fishery between them

The two counts are satisfied at the same time, with trailers per bank angler at
essentially zero:

$$
egin{aligned}
\lambda_{	ext{bank}} R_{V1} + \lambda_{	ext{boat}} R_{V2} &= V_I / b_1\
\lambda_{	ext{bank}} R_{T1} + \lambda_{	ext{boat}} R_{T2} &= T_I / b_2
\end{aligned}
$$

Writing $k_1 = b_1^{	ext{fitted}}/b_1$ and $k_2 = b_2^{	ext{fitted}}/b_2$, the
trailer equation pins boat effort and the vehicle equation then determines bank:

$$
rac{\lambda_{	ext{boat}}}{\lambda_{	ext{boat}}^{0}} = k_2,
\qquad
rac{\lambda_{	ext{bank}}}{\lambda_{	ext{bank}}^{0}} = k_1(1+ho) - ho\, k_2
$$

where $ho = (R_{V2}/R_{V1})\cdot(\lambda_{	ext{boat}}^{0}/\lambda_{	ext{bank}}^{0})$
is the boat-to-bank ratio of the vehicle count.

Two things follow that the simpler treatment misses. Moving $b_1$ alone does
**not** cleanly rescale the whole fishery, because boat effort is held in place
by the trailer counts, so the entire adjustment lands on bank. And moving $b_2$
alone does **not** leave bank untouched, because lowering boat effort frees
vehicle count to be attributed to bank, which pushes bank the other way. The two
partly offset each other.

Catch follows effort within each gear, since the catch rate is gear-specific and
carries no $b$.
:::

### The same range of `b`, in fish
)---")
```

```{r fish-table, eval = HAVE_FISH}
fish |>
  mutate(range = sprintf("%+.0f%% / %+.0f%%", pct_high, pct_low),
         yr = as.character(year_start)) |>
  select(fishery_type, yr, group, bias_type, fitted, if_b_low, if_b_high, range) |>
  gt(groupname_col = "fishery_type") |>
  cols_label(yr = "Year", group = "Catch group", bias_type = "b term",
             fitted = "As fitted",
             if_b_low = "If b is at the low end",
             if_b_high = "If b is at the high end",
             range = "Change") |>
  fmt_number(c(fitted, if_b_low, if_b_high), decimals = 0) |>
  cols_align("right", c(fitted, if_b_low, if_b_high, range)) |>
  sub_missing(everything(), missing_text = "--") |>
  tab_options(table.font.size = 13, data_row.padding = px(4),
              row_group.font.weight = "bold") |>
  tab_source_note(
    paste0("Fish. Vehicle rows move the whole season total; trailer rows move ",
           "the boat-caught portion only, which is why their percent change is ",
           "the smaller of the two wherever the fishery is mostly bank.")
  )
```

```{r fish-figure, eval = HAVE_FISH, fig.width = 9, fig.height = 6}
fish |>
  ggplot(aes(y = fct_rev(factor(fishery_name)), color = bias_type)) +
  geom_linerange(aes(xmin = if_b_high, xmax = if_b_low),
                 position = position_dodge(width = 0.55), linewidth = 2.2, alpha = 0.9) +
  # group = bias_type explicitly: setting color outside aes() strips the only
  # discrete aesthetic from this layer, so position_dodge() has nothing to
  # dodge by and the white halo lands on the row centre while the coloured
  # point it is supposed to sit under is offset.
  geom_point(aes(x = fitted, group = bias_type), position = position_dodge(width = 0.55),
             size = 2.6, color = SURFACE) +
  geom_point(aes(x = fitted), position = position_dodge(width = 0.55), size = 1.8) +
  facet_wrap(~group, scales = "free_x") +
  scale_x_continuous(labels = scales::comma) +
  scale_color_manual(values = c(vehicle = CAT[["blue"]], trailer = CAT[["orange"]]),
                     name = NULL) +
  labs(title = "Estimated catch under the plausible range of b", x = "Fish", y = NULL) +
  theme_bss()
```

```{r fish-figure-note, eval = HAVE_FISH, results = "asis"}
cat(r"---(
The point is the estimate as fitted; the bar spans what it would be at the two
ends of the plausible range for `b`.
)---")
```

