---
title: "Effort-index bias (`b`) and estimated catch"
subtitle: "Skagit / Snohomish / Stillaguamish freshwater creel"
date: today
format:
  html:
    embed-resources: true
    toc: false
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(gt)
library(here)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")
source(here::here("analysis", "bss_bias", "common.R"))

need <- function(f) {
  p <- file.path(OUT_DIR, f)
  if (!file.exists(p)) stop(f, " not found -- run 07_catch_sensitivity.R first.", call. = FALSE)
  readr::read_csv(p, show_col_types = FALSE)
}

T7  <- need("bss_b_T7_direct_sensitivity.csv")
cal <- need("bss_b_T6_calibration.csv")

# est_cg is the full species_lifestage_finmark_fate string. Collaborators do
# not need to read a regex, so it is labelled by target species.
group_label <- function(x) {
  dplyr::case_when(
    stringr::str_starts(x, "Chinook") ~ "Chinook encounters",
    stringr::str_starts(x, "Coho")    ~ "Coho harvest",
    is.na(x)                          ~ NA_character_,
    TRUE                              ~ stringr::str_extract(x, "^[^_]+")
  )
}

# One row per fishery-year x b term x catch group: what the fit produced, and
# what it would have been at the low and high ends of that series' own
# plausible range for b.
wide <- T7 |>
  filter(tier_kind %in% c("anchor", "empirical")) |>
  mutate(group = group_label(est_cg)) |>
  # pct_change_catch_total, not pct_change_catch: the first is the change in
  # the whole fishery's catch, which is what a reader quotes. They differ for
  # trailer, where only the boat component moves.
  # Pivot on a short stable key, not the display label: the labels are written
  # for the reader and will be reworded, and column names should not move when
  # they are.
  mutate(tier_kind_label = case_when(
    tier_kind == "anchor"                    ~ "fitted",
    str_detect(tier, "^low end")             ~ "low",
    str_detect(tier, "^high end")            ~ "high",
    TRUE                                     ~ NA_character_
  )) |>
  filter(!is.na(tier_kind_label)) |>
  select(fishery_name, year_start, bias_type, group, tier_kind_label,
         catch_estimate, pct_change_catch = pct_change_catch_total) |>
  pivot_wider(names_from = tier_kind_label, values_from = c(catch_estimate, pct_change_catch)) |>
  transmute(
    fishery_name, year_start, bias_type, group,
    fitted    = catch_estimate_fitted,
    if_b_low  = catch_estimate_low,
    if_b_high = catch_estimate_high,
    pct_low   = pct_change_catch_low,
    pct_high  = pct_change_catch_high
  ) |>
  filter(!is.na(fitted)) |>
  arrange(fishery_name, group, bias_type)
```

Estimated catch scales inversely with `b`: a higher `b` means the index
over-counts relative to anglers, so effort and catch are revised **down**, and
a lower `b` revises them **up**. Effort moves by exactly the same factor, so the
catch **rate** does not change — only the scale of the fishery.

::: {.callout-note collapse="true"}
## Why catch scales as 1/`b`

Three observations enter the effort side of the model, and only one carries
$b$ (notation as in the full report):

$$
\begin{aligned}
E_s &\sim \text{Poisson}\big(\lambda\big)
  &&\text{census (tie-in): everything in the block}\\[2pt]
V_I &\sim \text{Poisson}\big(\lambda \cdot R_V \cdot b_1\big)
  &&\text{vehicle index: the counted sites}\\[2pt]
V_A &\sim \text{Binomial}\big(A_A,\; R_V\big)
  &&\text{interviews: } A_A \text{ anglers, } V_A \text{ vehicles}
\end{aligned}
$$

**With no census**, the first line is absent and $\lambda$ — anglers present —
is pinned only by the index counts:

$$
\lambda \;=\; \frac{V_I}{R_V \;\cdot\; b_1}
$$

Catch is effort $\times$ trip length $\times$ catch rate, and the catch rate
$c$ comes from the interviews alone — it never sees $b$:

$$
C \;=\; \lambda \cdot L \cdot c
\;=\; \frac{V_I \cdot L \cdot c}{R_V \cdot b_1}
$$

$V_I$ is the count itself, $L$ is day length, and $R_V$ and $c$ both come from
the interviews. Every term but $b_1$ is fixed by something other than $b$, so

$$
C \;\propto\; \frac{1}{b_1}
$$

Effort is $E = \lambda \cdot L$, which carries the same $1/b_1$, so the catch
rate $C/E = c$ is **unchanged** by $b$. That is why `b` rescales the fishery
without changing the rate at which fish are caught.

**With census**, $E_s \sim \text{Poisson}(\lambda)$ pins $\lambda$ directly,
$b$ no longer has to absorb the difference, and the effect is smaller than
shown here. The tables below are therefore the census-free case — the scenario
under discussion, and the larger of the two.
:::

Using a predicted `b` instead of measuring one with census counts moves the
catch estimate by the amounts below. Low and high are the ends of the **95% prediction interval** for `b` in that
series: the range a new, unmeasured year is expected to fall in, estimated from
the other years of that same fishery. Not hypothetical values.

```{r calibration-line, results = "asis"}
v <- cal |> filter(bias_type == "vehicle")
t <- cal |> filter(bias_type == "trailer")
cat(sprintf(
  paste0("Across %d held-out fishery-years, a predicted **vehicle** `b` would have been within ",
         "**%.0f%%** of the measured value in the typical year (worst case %.0f%%). The ",
         "**trailer** term is considerably less reliable: typically **%.0f%%**, worst case ",
         "**%.0f%%**.\n"),
  sum(cal$n_years_tested),
  median(v$median_abs_pct_err), max(abs(v$worst_pct_err)),
  median(t$median_abs_pct_err), max(abs(t$worst_pct_err))
))
```

## What changing `b` does to estimated catch

```{r table}
wide |>
  mutate(range = sprintf("%+.0f%% / %+.0f%%", pct_high, pct_low)) |>
  select(fishery_name, group, bias_type, fitted, if_b_low, if_b_high, range) |>
  gt(groupname_col = "fishery_name") |>
  cols_label(
    group     = "Catch group",
    bias_type = "b term",
    fitted    = "As fitted",
    if_b_low  = "Low end of predicted b",
    if_b_high = "High end of predicted b",
    range     = "Range"
  ) |>
  fmt_number(c(fitted, if_b_low, if_b_high), decimals = 0) |>
  cols_align("right", c(fitted, if_b_low, if_b_high, range)) |>
  tab_options(table.font.size = 13, data_row.padding = px(4),
              row_group.font.weight = "bold") |>
  tab_source_note(
    "Fish. Low and high are the ends of the 95% prediction interval for b in that series -- the range a new, unmeasured year is expected to fall in, estimated from the other years of that fishery."
  )
```

## Catch and effort by gear

The vehicle index counts every angler's vehicle; the trailer index counts boat
trailers. So the two terms act on different parts of the fishery, and the split
between bank and boat is what sets how far each one reaches.

```{r gear-split}
t8_path <- file.path(OUT_DIR, "bss_b_T8_gear_split.csv")
if (file.exists(t8_path)) {
  readr::read_csv(t8_path, show_col_types = FALSE) |>
    filter(tier_kind == "anchor", bias_type == "vehicle") |>
    mutate(group = group_label(est_cg)) |>
    select(fishery_name, group, catch_bank, catch_boat, effort_bank, effort_boat) |>
    gt() |>
    cols_label(fishery_name = "Fishery", group = "Catch group",
               catch_bank = "Bank", catch_boat = "Boat",
               effort_bank = "Bank", effort_boat = "Boat") |>
    tab_spanner("Catch (fish)", c(catch_bank, catch_boat)) |>
    tab_spanner("Effort (angler-hours)", c(effort_bank, effort_boat)) |>
    fmt_number(c(catch_bank, catch_boat, effort_bank, effort_boat), decimals = 0) |>
    sub_missing(everything(), missing_text = "--") |>
    tab_options(table.font.size = 13, data_row.padding = px(4))
} else {
  cat("*Gear split not available -- run 09_read_production_estimates.R, then 07.*")
}
```

```{r figure, fig.width = 9, fig.height = 6}
fig <- wide |>
  mutate(label = paste0(fishery_name)) |>
  ggplot(aes(y = fct_rev(factor(label)), color = bias_type)) +
  geom_linerange(aes(xmin = if_b_high, xmax = if_b_low),
                 position = position_dodge(width = 0.55), linewidth = 2.2, alpha = 0.9) +
  geom_point(aes(x = fitted), position = position_dodge(width = 0.55),
             size = 2.6, color = SURFACE) +
  geom_point(aes(x = fitted), position = position_dodge(width = 0.55), size = 1.8) +
  facet_wrap(~group, scales = "free_x") +
  scale_x_continuous(labels = scales::comma) +
  scale_color_manual(values = c(vehicle = CAT[["blue"]], trailer = CAT[["orange"]]), name = NULL) +
  labs(title = "Estimated catch under the plausible range of b", x = "Fish", y = NULL) +
  theme_bss()

fig
```

The point is the estimate as fitted; the bar spans what it would be at the ends
of that series' plausible range for `b`.

---

*Method: `b` is the bias term on the effort index count in the BSS creel model;
the derivation above is the one in the full report. Reliability figures are a leave-one-out
backtest: each year predicted from the other years of its series, never from a
mean containing itself. Full detail and code in `analysis/bss_bias/`.*
