---
title: "Effort-index bias (`b`): what it does to estimated catch"
subtitle: "Skagit / Snohomish / Stillaguamish freshwater creel -- part 1 of 2"
date: today
format:
  html:
    embed-resources: true
    toc: false
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# ==============================================================================
# PART 1 -- the mechanism and the reliability of a borrowed b.
#
# DELIBERATELY SELF-CONTAINED ON THE b DRAWS. This document reads T7 and the T6
# calibration only, and touches NO season totals: not bss_catch_baseline.csv,
# not catch_estimate, not the gear split. Everything here comes from the b
# posteriors and the meta-analysis, both of which are complete.
#
# That is the point of the split. Part 1 can be sent without waiting on a model
# sweep, and a stale or half-finished baseline sitting in outputs/ cannot leak
# into it. Season totals in fish, and the bank/boat split, are part 2
# (12_estimates_update.qmd) -- the complementary update, not a correction.
# ==============================================================================
library(tidyverse)
library(gt)
library(here)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")
source(here::here("analysis", "bss_bias", "common.R"))

need <- function(f) {
  p <- file.path(OUT_DIR, f)
  if (!file.exists(p)) stop(f, " not found -- run 07_catch_sensitivity.R first.", call. = FALSE)
  readr::read_csv(p, show_col_types = FALSE)
}

T7  <- need("bss_b_T7_direct_sensitivity.csv")
cal <- need("bss_b_T6_calibration.csv")

# One row per fishery-year x b term: b as fitted, the ends of that series' own
# plausible range, and the percent change in catch at each end.
#
# pct_change_catch, NOT pct_change_catch_total: the first is 100*(b_fitted/b_alt - 1),
# which needs no season total. The second divides by C_sum and is NA until the
# fits land -- it belongs to part 2.
#
# Pivot on a short stable key rather than the display label, so rewording a
# label for the reader does not move a column name.
wide <- T7 |>
  filter(tier_kind %in% c("anchor", "empirical")) |>
  mutate(key = case_when(
    tier_kind == "anchor"        ~ "fitted",
    str_detect(tier, "^low end") ~ "low",
    str_detect(tier, "^high end")~ "high",
    TRUE                         ~ NA_character_
  )) |>
  filter(!is.na(key)) |>
  distinct(fishery_type, fishery_name, year_start, bias_type, key,
           b_alt, pct_change_catch) |>
  pivot_wider(names_from = key, values_from = c(b_alt, pct_change_catch)) |>
  transmute(
    fishery_type, fishery_name, year_start, bias_type,
    b_fitted = b_alt_fitted,
    b_low    = b_alt_low,
    b_high   = b_alt_high,
    pct_low  = pct_change_catch_low,     # b at the LOW end -> catch UP
    pct_high = pct_change_catch_high     # b at the HIGH end -> catch DOWN
  ) |>
  filter(!is.na(b_low), !is.na(b_high)) |>
  arrange(fishery_type, year_start, bias_type)
```

`b` is the correction between what an index count sees and how many anglers are
actually out there. Measuring it needs census (tie-in) counts. The question on
the table is whether a `b` carried over from previous years can stand in for a
measured one in a year where census counts are not collected -- and if it can,
how much the catch estimate moves when that borrowed `b` is off.

**Catch moves one-for-one in the opposite direction from `b`.** Double `b` and
the catch estimate is cut in half; halve `b` and it doubles; a `b` that is 10%
too high pulls catch 10% too low. Estimated effort moves by exactly the same
factor, so the catch **rate** -- fish per angler-hour -- does not change at all.
`b` rescales the size of the fishery, not how well it is fishing.

::: {.callout-note collapse="true"}
## Why catch is proportional to 1/`b`

Three observations enter the effort side of the model, and only one carries
$b$ (notation as in the full report):

$$
\begin{aligned}
E_s &\sim \text{Poisson}\big(\lambda\big)
  &&\text{census (tie-in): everything in the block}\\[2pt]
V_I &\sim \text{Poisson}\big(\lambda \cdot R_V \cdot b_1\big)
  &&\text{vehicle index: the counted sites}\\[2pt]
V_A &\sim \text{Binomial}\big(A_A,\; R_V\big)
  &&\text{interviews: } A_A \text{ anglers, } V_A \text{ vehicles}
\end{aligned}
$$

**With no census**, the first line is absent and $\lambda$ -- anglers present --
is pinned only by the index counts:

$$
\lambda \;=\; \frac{V_I}{R_V \;\cdot\; b_1}
$$

Catch is effort $\times$ trip length $\times$ catch rate, and the catch rate
$c$ comes from the interviews alone -- it never sees $b$:

$$
C \;=\; \lambda \cdot L \cdot c
\;=\; \frac{V_I \cdot L \cdot c}{R_V \cdot b_1}
$$

$V_I$ is the count itself, $L$ is day length, and $R_V$ and $c$ both come from
the interviews. Every term but $b_1$ is fixed by something other than $b$, so

$$
C \;\propto\; \frac{1}{b_1}
$$

Effort is $E = \lambda \cdot L$, which carries the same $1/b_1$, so the catch
rate $C/E = c$ is **unchanged** by $b$. That is why `b` rescales the fishery
without changing the rate at which fish are caught.

**With census**, $E_s \sim \text{Poisson}(\lambda)$ pins $\lambda$ directly,
$b$ no longer has to absorb the difference, and the effect is smaller than
shown here. The tables below are therefore the census-free case -- the scenario
under discussion, and the larger of the two.
:::

## How close a borrowed `b` gets

```{r calibration-line, results = "asis"}
v <- cal |> filter(bias_type == "vehicle")
t <- cal |> filter(bias_type == "trailer")
cat(sprintf(
  paste0("Each fishery-year with a measured `b` was predicted from the **other** years of ",
         "its own series and nothing else, then compared against what was actually ",
         "measured -- %d such tests in all. A predicted **vehicle** `b` landed within ",
         "**%.0f%%** of the measured value in the typical year, worst case %.0f%%. The ",
         "**trailer** term is considerably less reliable: typically **%.0f%%** off, worst ",
         "case **%.0f%%**.\n"),
  sum(cal$n_years_tested),
  median(v$median_abs_pct_err), max(abs(v$worst_pct_err)),
  median(t$median_abs_pct_err), max(abs(t$worst_pct_err))
))
```

Because catch tracks `b` one-for-one, those are also the errors carried straight
into the catch estimate. **The vehicle term is accurate enough to borrow; the
trailer term is not.**

## What the plausible range of `b` does to estimated catch

Low and high below are the ends of the **95% prediction interval** for `b` in
that series -- the range a new, unmeasured year is expected to fall in, worked
out from the other years of that same fishery. They are not hypothetical round
numbers; they are what `b` could actually be here.

```{r table}
# The two index channels do not observe the same anglers. The vehicle count sees
# every angler's vehicle, so b[1] rescales the whole fishery. Trailers per BANK
# angler go to ~0 in the interviews, so the trailer count sees boat anglers
# essentially alone and b[2] moves the boat-caught portion. Saying so in the
# row label keeps the percent honest without needing the gear totals, which is
# exactly what part 2 supplies.
wide |>
  mutate(
    term = if_else(bias_type == "vehicle",
                   "vehicle -- whole fishery",
                   "trailer -- boat-caught portion"),
    b_range = sprintf("%.2f to %.2f", b_low, b_high),
    yr = as.character(year_start)
  ) |>
  select(fishery_type, yr, term, b_fitted, b_range, pct_low, pct_high) |>
  gt(groupname_col = "fishery_type") |>
  cols_label(yr = "Year", term = "b term", b_fitted = "b as fitted",
             b_range = "Plausible range for b",
             pct_low = "If b is at the low end",
             pct_high = "If b is at the high end") |>
  fmt_number(b_fitted, decimals = 2) |>
  fmt_number(c(pct_low, pct_high), decimals = 0, pattern = "{x}%", force_sign = TRUE) |>
  cols_align("right", c(b_fitted, b_range, pct_low, pct_high)) |>
  sub_missing(everything(), missing_text = "--") |>
  tab_options(table.font.size = 13, data_row.padding = px(4),
              row_group.font.weight = "bold") |>
  tab_source_note(
    paste0("Change in estimated catch. The percentage is the same for every species and ",
           "catch group, because b moves effort and leaves the catch rate alone. ",
           "Vehicle rows move the whole fishery; trailer rows move the boat-caught ",
           "portion, so their effect on the season total is smaller by however much of ",
           "the fishery is fished from the bank.")
  )
```

```{r figure, fig.width = 9, fig.height = 6}
wide |>
  mutate(term = if_else(bias_type == "vehicle",
                        "vehicle (whole fishery)", "trailer (boat-caught portion)")) |>
  ggplot(aes(y = fct_rev(factor(fishery_name)), color = term)) +
  geom_vline(xintercept = 0, color = INK_SECOND, linewidth = 0.5, linetype = "22") +
  geom_linerange(aes(xmin = pct_high, xmax = pct_low),
                 position = position_dodge(width = 0.55), linewidth = 2.2, alpha = 0.9) +
  scale_x_continuous(labels = function(x) paste0(ifelse(x > 0, "+", ""), x, "%")) +
  scale_color_manual(values = c("vehicle (whole fishery)" = CAT[["blue"]],
                                "trailer (boat-caught portion)" = CAT[["orange"]]),
                     name = NULL) +
  labs(title = "Change in estimated catch across the plausible range of b",
       x = "Change in estimated catch", y = NULL) +
  theme_bss()
```

Each bar spans what the catch estimate would be at the two ends of the plausible
range for `b`. The right-hand end is the high `b` (catch revised down), the
left-hand end the low `b` (catch revised up).

## What is not in this part

Season totals in fish, and the split between bank and boat, are being fitted
now and follow in **part 2**. They do not change anything above -- the
percentages hold for every species and catch group -- but they put the effect in
fish, and they show how much of each fishery the trailer term can actually
reach.

---

*Method: `b` is the bias term on the effort index count in the BSS creel model;
the derivation above is the one in the full report. Reliability figures are a
leave-one-out backtest -- each year predicted from the other years of its series,
never from an average containing itself. The catch effects shown are an upper
bound: they assume no census counts at all, which is the scenario under
discussion. In a year that does collect census, the census partly corrects an
imported `b` and the effect is smaller. Full detail and code in
`analysis/bss_bias/`.*
