---
title: "Effort-index bias (`b`): what it does to estimated catch"
date: today
params:
  focus: "Stillaguamish"
format:
  html:
    embed-resources: true
    toc: false
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# ==============================================================================
# PART 1 -- the mechanism and the reliability of a borrowed b.
#
# DELIBERATELY SELF-CONTAINED ON THE b DRAWS. This document reads T7 and the T6
# calibration only, and touches NO season totals: not bss_catch_baseline.csv,
# not catch_estimate, not the gear split. Everything here comes from the b
# posteriors and the meta-analysis, both of which are complete.
#
# That is the point of the split. Part 1 can be sent without waiting on a model
# sweep, and a stale or half-finished baseline sitting in outputs/ cannot leak
# into it. Season totals in fish, and the bank/boat split, are part 2
# (12_estimates_update.qmd) -- the complementary update, not a correction.
# ==============================================================================
library(tidyverse)
library(gt)
library(here)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")
source(here::here("analysis", "bss_bias", "common.R"))

need <- function(f) {
  p <- file.path(OUT_DIR, f)
  if (!file.exists(p)) stop(f, " not found -- run 07_catch_sensitivity.R first.", call. = FALSE)
  readr::read_csv(p, show_col_types = FALSE)
}

T7  <- need("bss_b_T7_direct_sensitivity.csv")
cal <- need("bss_b_T6_calibration.csv")

# Which series this brief is about. A PARAM, not a constant, so one document
# renders a per-basin brief for each collaborator group:
#
#   quarto render 11_share_brief.qmd -P focus:Stillaguamish ...
#   quarto render 11_share_brief.qmd -P focus:Snohomish ...
#
# NOTE what is NOT subset by it: the backtest below. How well a borrowed b
# predicts is a question about the METHOD, and any one basin contributes only a
# handful of the 43 held-out tests. Scoring the method on those alone would be a
# weaker claim resting on a smaller sample, so the reliability figures stay
# across all three basins and the focal basin's own rows are called out within
# them.
FOCUS_RE <- params$focus

# One row per fishery-year x b term: b as fitted, the ends of that series' own
# plausible range, and the percent change in catch at each end.
#
# pct_change_catch, NOT pct_change_catch_total: the first is 100*(b_fitted/b_alt - 1),
# which needs no season total. The second divides by C_sum and is NA until the
# fits land -- it belongs to part 2.
#
# Pivot on a short stable key rather than the display label, so rewording a
# label for the reader does not move a column name.
wide <- T7 |>
  filter(tier_kind %in% c("anchor", "empirical")) |>
  mutate(key = case_when(
    tier_kind == "anchor"        ~ "fitted",
    str_detect(tier, "^low end") ~ "low",
    str_detect(tier, "^high end")~ "high",
    TRUE                         ~ NA_character_
  )) |>
  filter(!is.na(key)) |>
  distinct(fishery_type, fishery_name, year_start, bias_type, key,
           b_alt, pct_change_catch) |>
  pivot_wider(names_from = key, values_from = c(b_alt, pct_change_catch)) |>
  transmute(
    fishery_type, fishery_name, year_start, bias_type,
    b_fitted = b_alt_fitted,
    b_low    = b_alt_low,
    b_high   = b_alt_high,
    pct_low  = pct_change_catch_low,     # b at the LOW end -> catch UP
    pct_high = pct_change_catch_high     # b at the HIGH end -> catch DOWN
  ) |>
  filter(!is.na(b_low), !is.na(b_high)) |>
  filter(str_detect(fishery_type, FOCUS_RE)) |>
  arrange(fishery_type, year_start, bias_type)

if (nrow(wide) == 0) stop("No fishery_type matches focus = '", FOCUS_RE, "'.", call. = FALSE)

# The series name for prose and headings, taken from the data rather than passed
# in. Guarantees the label matches what the filter actually kept, and keeps the
# only command-line param a single word with no quoting hazard.
SERIES_LABEL <- {
  ft <- sort(unique(wide$fishery_type))
  if (length(ft) == 1) ft else paste(ft, collapse = " / ")
}

```

`b` is the correction between what an index count sees and how many anglers are
actually out there. Measuring it needs census (tie-in) counts. Our key question
is whether a `b` carried over from previous years can stand in for a
measured one in a year where census counts are not collected, and if it can,
how much the catch estimate moves due to changes in `b`.

::: {.callout-note collapse="true"}
## Why catch is proportional to 1/`b`

Three observations enter the effort side of the model, and only one carries
$b$ (notation as in the full report):

$$
\begin{aligned}
E_s &\sim \text{Poisson}\big(\lambda\big)
  &&\text{census (tie-in): everything in the block}\\[2pt]
V_I &\sim \text{Poisson}\big(\lambda \cdot R_V \cdot b_1\big)
  &&\text{vehicle index: the counted sites}\\[2pt]
V_A &\sim \text{Binomial}\big(A_A,\; R_V\big)
  &&\text{interviews: } A_A \text{ anglers, } V_A \text{ vehicles}
\end{aligned}
$$

**With no census**, the first line is absent and $\lambda$ -- anglers present --
is pinned only by the index counts:

$$
\lambda \;=\; \frac{V_I}{R_V \;\cdot\; b_1}
$$

Catch is effort $\times$ trip length $\times$ catch rate, and the catch rate
$c$ comes from the interviews alone -- it never sees $b$:

$$
C \;=\; \lambda \cdot L \cdot c
\;=\; \frac{V_I \cdot L \cdot c}{R_V \cdot b_1}
$$

$V_I$ is the count itself, $L$ is day length, and $R_V$ and $c$ both come from
the interviews. Every term but $b_1$ is fixed by something other than $b$, so

$$
C \;\propto\; \frac{1}{b_1}
$$

Effort is $E = \lambda \cdot L$, which carries the same $1/b_1$, so the catch
rate $C/E = c$ is **unchanged** by $b$. That is why `b` rescales the fishery
without changing the rate at which fish are caught.

:::

## What the plausible range of `b` does to estimated catch

Low and high below are the ends of the **95% prediction interval** for `b` in
that series, based on an earlier part of this analysis.
This can be interpreted as the range a new, unmeasured year could be expected to fall in, derived from
other years of that same fishery. Essentially, they are what `b` could be given what has been observed
in the past.

```{r table}
# The two index channels do not observe the same anglers. The vehicle count sees
# every angler's vehicle, so b[1] rescales the whole fishery. Trailers per BANK
# angler go to ~0 in the interviews, so the trailer count sees boat anglers
# essentially alone and b[2] moves the boat-caught portion. Saying so in the
# row label keeps the percent honest without needing the gear totals, which is
# exactly what part 2 supplies.
wide |>
  mutate(
    term = if_else(bias_type == "vehicle",
                   "vehicle -- whole fishery",
                   "trailer -- boat-caught portion"),
    b_range = sprintf("%.2f to %.2f", b_low, b_high),
    yr = as.character(year_start)
  ) |>
  select(fishery_type, yr, term, b_fitted, b_range, pct_low, pct_high) |>
  gt(groupname_col = "fishery_type") |>
  cols_label(yr = "Year", term = "b term", b_fitted = "b as fitted",
             b_range = "Plausible range for b",
             pct_low = "If b is at the low end",
             pct_high = "If b is at the high end") |>
  fmt_number(b_fitted, decimals = 2) |>
  fmt_number(c(pct_low, pct_high), decimals = 0, pattern = "{x}%", force_sign = TRUE) |>
  cols_align("right", c(b_fitted, b_range, pct_low, pct_high)) |>
  sub_missing(everything(), missing_text = "--") |>
  tab_options(table.font.size = 13, data_row.padding = px(4),
              row_group.font.weight = "bold") |>
  tab_source_note(
    paste0("Change in estimated catch. The percentage is the same for every species and ",
           "catch group, because b moves effort and leaves the catch rate alone. ",
           "Vehicle rows move the whole fishery; trailer rows move the boat-caught ",
           "portion, so their effect on the season total differs by however much of ",
           "the fishery is fished from the bank.")
  )
```

```{r figure, fig.width = 9, fig.height = 6}
wide |>
  mutate(term = if_else(bias_type == "vehicle",
                        "vehicle (whole fishery)", "trailer (boat-caught portion)")) |>
  ggplot(aes(y = fct_rev(factor(fishery_name)), color = term)) +
  geom_vline(xintercept = 0, color = INK_SECOND, linewidth = 0.5, linetype = "22") +
  geom_linerange(aes(xmin = pct_high, xmax = pct_low),
                 position = position_dodge(width = 0.55), linewidth = 2.2, alpha = 0.9) +
  scale_x_continuous(labels = function(x) paste0(ifelse(x > 0, "+", ""), x, "%")) +
  scale_color_manual(values = c("vehicle (whole fishery)" = CAT[["blue"]],
                                "trailer (boat-caught portion)" = CAT[["orange"]]),
                     name = NULL) +
  labs(title = "Change in estimated catch across the plausible range of b",
       x = "Change in estimated catch", y = NULL) +
  theme_bss()
```

Each bar spans what the catch estimate would be at the two ends of the plausible
range for `b`. The right-hand end is the high `b` (effort, and thus catch, revised down), the
left-hand end the low `b` (effort, and thus, catch revised up).

## How close a borrowed `b` gets

```{r calibration-line, results = "asis"}
# Deliberately NOT subset to FOCUS_RE -- see the note in the setup chunk. How
# well the method predicts is scored on every series, then the focal basin's own
# rows are reported inside that.
v <- cal |> filter(bias_type == "vehicle")
t <- cal |> filter(bias_type == "trailer")
cat(sprintf(
  paste0("Each fishery-year with a measured `b` was predicted from the **other** years of ",
         "its own series and nothing else, then compared against what was actually ",
         "measured -- %d such tests across the three basins. A predicted **vehicle** `b` ",
         "landed within **%.0f%%** of the measured value in the typical year, worst case ",
         "%.0f%%. The **trailer** term is less reliable: typically ",
         "**%.0f%%** off, worst case **%.0f%%**.\n\n"),
  sum(cal$n_years_tested),
  median(v$median_abs_pct_err), max(abs(v$worst_pct_err)),
  median(t$median_abs_pct_err), max(abs(t$worst_pct_err))
))

foc <- cal |> filter(str_detect(fishery_type, FOCUS_RE))
fv  <- foc |> filter(bias_type == "vehicle")
ft  <- foc |> filter(bias_type == "trailer")
# A basin missing one of the two terms would otherwise print NA and -Inf into a
# document going to co-managers. Say nothing rather than that.
if (nrow(fv) > 0 && nrow(ft) > 0) cat(sprintf(
  paste0("%s contributes %d of those tests. Its **vehicle** term came in ",
         "inside the interval %d of %d times, typically **%.0f%%** off the measured ",
         "value (worst %.0f%%); its **trailer** term %d of %d, typically **%.0f%%** off ",
         "(worst %.0f%%).\n"),
  SERIES_LABEL, sum(foc$n_years_tested),
  sum(fv$n_inside_95_pi), sum(fv$n_years_tested),
  median(fv$median_abs_pct_err), max(abs(fv$worst_pct_err)),
  sum(ft$n_inside_95_pi), sum(ft$n_years_tested),
  median(ft$median_abs_pct_err), max(abs(ft$worst_pct_err))
))
```

## Follow up

Season total catch estimates by bank and boat gear strata, are being fitted
overnight. These updated results will put the effect in
fish, and will also help illustrate the magnitude of bank/boat angling effort and contextualize
their respective b terms.

