---
title: "Estimated effort and catch by gear, and what `b` moves"
subtitle: "Skagit / Snohomish / Stillaguamish freshwater creel -- part 2 of 2"
date: today
format:
  html:
    embed-resources: true
    toc: false
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
# ==============================================================================
# PART 2 -- the same result in fish, plus the bank/boat split.
#
# COMPLEMENT TO 11_share_brief.qmd, NOT A REPLACEMENT. Part 1 stands on the b
# draws alone and needs no season totals; everything it says still holds. This
# document adds what the fitted season totals make possible: the effect in fish
# rather than percent, and the gear split that decides how far the trailer term
# can actually reach.
#
# REQUIRES the model sweep to have finished:
#   Rscript analysis/bss_bias/09_read_production_estimates.R
#   Rscript analysis/bss_bias/07_catch_sensitivity.R
# The stop() below is deliberate. Rendering this against a stale or partial
# bss_catch_baseline.csv is how bad season totals reach a co-manager, so it
# fails loudly rather than printing whatever happens to be on disk.
# ==============================================================================
library(tidyverse)
library(gt)
library(here)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")
source(here::here("analysis", "bss_bias", "common.R"))

need <- function(f) {
  p <- file.path(OUT_DIR, f)
  if (!file.exists(p)) stop(f, " not found -- run 09_read_production_estimates.R then ",
                            "07_catch_sensitivity.R first.", call. = FALSE)
  readr::read_csv(p, show_col_types = FALSE)
}

T7 <- need("bss_b_T7_direct_sensitivity.csv")
T8 <- need("bss_b_T8_gear_split.csv")

if (all(is.na(T7$catch_estimate))) {
  stop("T7 has no season totals -- bss_catch_baseline.csv was missing when ",
       "07_catch_sensitivity.R ran. Run 09_read_production_estimates.R, then 07, ",
       "then re-render. Part 1 (11_share_brief.qmd) does not need this.",
       call. = FALSE)
}

# est_cg is the full species_lifestage_finmark_fate string. Collaborators do not
# need to read a regex, so it is labelled by target species.
group_label <- function(x) {
  dplyr::case_when(
    stringr::str_starts(x, "Chinook") ~ "Chinook encounters",
    stringr::str_starts(x, "Coho")    ~ "Coho harvest",
    is.na(x)                          ~ NA_character_,
    TRUE                              ~ stringr::str_extract(x, "^[^_]+")
  )
}

# At the anchor tier b_alt == b_fitted, so both gear factors are 1 and the row
# is the fit's own season total. Either bias_type gives the same numbers there;
# vehicle is taken arbitrarily.
as_fitted <- T8 |>
  filter(tier_kind == "anchor", bias_type == "vehicle") |>
  mutate(group = group_label(est_cg),
         boat_share_catch  = catch_boat / catch_total,
         boat_share_effort = effort_boat / effort_total) |>
  arrange(fishery_name, group)

# One row per fishery-year x catch group x b term: the fit's total, and the
# total at each end of that series' plausible range for b.
wide <- T7 |>
  filter(tier_kind %in% c("anchor", "empirical")) |>
  mutate(group = group_label(est_cg),
         key = case_when(
           tier_kind == "anchor"         ~ "fitted",
           str_detect(tier, "^low end")  ~ "low",
           str_detect(tier, "^high end") ~ "high",
           TRUE                          ~ NA_character_
         )) |>
  filter(!is.na(key)) |>
  select(fishery_type, fishery_name, year_start, bias_type, group, key,
         catch_estimate, pct_change_catch_total) |>
  pivot_wider(names_from = key,
              values_from = c(catch_estimate, pct_change_catch_total)) |>
  transmute(
    fishery_type, fishery_name, year_start, bias_type, group,
    fitted    = catch_estimate_fitted,
    if_b_low  = catch_estimate_low,
    if_b_high = catch_estimate_high,
    pct_low   = pct_change_catch_total_low,
    pct_high  = pct_change_catch_total_high
  ) |>
  filter(!is.na(fitted)) |>
  arrange(fishery_type, year_start, group, bias_type)
```

Part 1 established the relationship: **estimated catch moves one-for-one in the
opposite direction from `b`**, and the catch rate does not move at all. The
percentages there hold for every species and catch group. This part puts the
same result in fish, and splits it by gear.

## Season totals as fitted

```{r fitted-totals}
as_fitted |>
  select(fishery_name, group, catch_bank, catch_boat, catch_total,
         effort_bank, effort_boat, boat_share_catch) |>
  gt(groupname_col = "fishery_name") |>
  cols_label(group = "Catch group",
             catch_bank = "Bank", catch_boat = "Boat", catch_total = "Total",
             effort_bank = "Bank", effort_boat = "Boat",
             boat_share_catch = "Boat share") |>
  tab_spanner("Catch (fish)", c(catch_bank, catch_boat, catch_total)) |>
  tab_spanner("Effort (angler-hours)", c(effort_bank, effort_boat)) |>
  fmt_number(c(catch_bank, catch_boat, catch_total), decimals = 0) |>
  fmt_number(c(effort_bank, effort_boat), decimals = 0) |>
  fmt_percent(boat_share_catch, decimals = 0) |>
  sub_missing(everything(), missing_text = "--") |>
  cols_align("right", c(catch_bank, catch_boat, catch_total,
                        effort_bank, effort_boat, boat_share_catch)) |>
  tab_options(table.font.size = 13, data_row.padding = px(4),
              row_group.font.weight = "bold") |>
  tab_source_note(
    paste0("Posterior medians. Gear totals are the model's own -- the daily, ",
           "section-level draws summed within gear, not a share applied afterwards.")
  )
```

## Why the two `b` terms reach different parts of the fishery

The vehicle index counts vehicles, which every angler arrives in, so `b[1]`
rescales the **whole fishery**. The trailer index counts boat trailers, and
trailers per bank angler come out at essentially zero in the interviews, so the
trailer count observes **boat anglers alone**.

That has a consequence worth stating plainly: in a fishery that is mostly fished
from the bank, the trailer term has very little of the fishery to move, and
getting it wrong costs correspondingly little. The boat share above is what sets
that. It is the direct answer to how much the trailer bias correction matters.

::: {.callout-note collapse="true"}
## The exact two-channel arithmetic

The two counts are satisfied at the same time, with $R_{T}$ for bank anglers at
essentially zero:

$$
\begin{aligned}
\lambda_{\text{bank}} R_{V1} + \lambda_{\text{boat}} R_{V2} &= V_I / b_1\\
\lambda_{\text{bank}} R_{T1} + \lambda_{\text{boat}} R_{T2} &= T_I / b_2
\end{aligned}
$$

Writing $k_1 = b_1^{\text{fitted}}/b_1$ and $k_2 = b_2^{\text{fitted}}/b_2$, the
trailer equation pins boat effort and the vehicle equation then determines bank:

$$
\frac{\lambda_{\text{boat}}}{\lambda_{\text{boat}}^{0}} = k_2,
\qquad
\frac{\lambda_{\text{bank}}}{\lambda_{\text{bank}}^{0}} = k_1(1+\rho) - \rho\, k_2
$$

where $\rho = (R_{V2}/R_{V1})\cdot(\lambda_{\text{boat}}^{0}/\lambda_{\text{bank}}^{0})$
is the boat-to-bank ratio of the vehicle count.

Two things fall out that the simpler treatment misses. Moving $b_1$ alone does
**not** cleanly rescale the whole fishery -- boat effort is held in place by the
trailer counts, so the entire adjustment lands on bank. And moving $b_2$ alone
does **not** leave bank untouched -- lowering boat effort frees vehicle count to
be attributed to bank, which pushes bank the other way. The two dials partly
offset each other.

Catch follows effort within each gear, because the catch rate is gear-specific
and carries no $b$.
:::

## Estimated catch across the plausible range of `b`

Low and high are the ends of the **95% prediction interval** for `b` in that
series -- the range a new, unmeasured year is expected to fall in, worked out
from the other years of that same fishery.

```{r sensitivity-table}
wide |>
  mutate(range = sprintf("%+.0f%% / %+.0f%%", pct_high, pct_low),
         yr = as.character(year_start)) |>
  select(fishery_type, yr, group, bias_type, fitted, if_b_low, if_b_high, range) |>
  gt(groupname_col = "fishery_type") |>
  cols_label(yr = "Year", group = "Catch group", bias_type = "b term",
             fitted = "As fitted",
             if_b_low = "If b is at the low end",
             if_b_high = "If b is at the high end",
             range = "Change") |>
  fmt_number(c(fitted, if_b_low, if_b_high), decimals = 0) |>
  cols_align("right", c(fitted, if_b_low, if_b_high, range)) |>
  sub_missing(everything(), missing_text = "--") |>
  tab_options(table.font.size = 13, data_row.padding = px(4),
              row_group.font.weight = "bold") |>
  tab_source_note(
    paste0("Fish. Vehicle rows move the whole season total; trailer rows move the ",
           "boat-caught portion only, which is why their percent change is the ",
           "smaller of the two wherever the fishery is mostly bank.")
  )
```

```{r figure, fig.width = 9, fig.height = 6}
wide |>
  ggplot(aes(y = fct_rev(factor(fishery_name)), color = bias_type)) +
  geom_linerange(aes(xmin = if_b_high, xmax = if_b_low),
                 position = position_dodge(width = 0.55), linewidth = 2.2, alpha = 0.9) +
  geom_point(aes(x = fitted), position = position_dodge(width = 0.55),
             size = 2.6, color = SURFACE) +
  geom_point(aes(x = fitted), position = position_dodge(width = 0.55), size = 1.8) +
  facet_wrap(~group, scales = "free_x") +
  scale_x_continuous(labels = scales::comma) +
  scale_color_manual(values = c(vehicle = CAT[["blue"]], trailer = CAT[["orange"]]),
                     name = NULL) +
  labs(title = "Estimated catch under the plausible range of b", x = "Fish", y = NULL) +
  theme_bss()
```

The point is the estimate as fitted; the bar spans what it would be at the two
ends of the plausible range for `b`.

---

*Method: as part 1. Season totals are posterior medians from the production BSS
fits, with gear totals summed from the daily section-level draws. The `b`
effects assume no census counts in the year being estimated, which is the upper
bound -- with census, the census partly corrects an imported `b` and the effect
is smaller. Full detail and code in `analysis/bss_bias/`.*
