---
title: "BSS effort-bias term (`b`): interannual variability and a predictive approach"
subtitle: "DFW/tribal technical staff working session"
date: last-modified
execute:
  warning: false
  message: false
  echo: false
format:
  html:
    page-layout: full
    code-tools: true
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    title-block-banner: "#1D886E"
---

```{r setup}
library(tidyverse)
library(here)
library(gt)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")
FIG_DIR <- file.path(OUT_DIR, "figures")

# This report is a PRESENTATION LAYER over analysis/bss_bias/outputs/ -- it
# reads CSVs and embeds figures that the numbered scripts already produced,
# and never recomputes anything. That is deliberate: it means the report
# cannot quietly disagree with the pipeline that generated it, it needs no
# database or VPN, and it re-renders in seconds after each refit.

read_out <- function(name) {
  p <- file.path(OUT_DIR, name)
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
}

fig_path <- function(name) {
  p <- file.path(FIG_DIR, paste0(name, ".png"))
  if (file.exists(p)) p else NULL
}

# Interactive where it earns its keep: tables are for interrogating (sort by
# b, filter to one river, search a fishery), so they get DT. Figures are for
# reading, so they are embedded as produced. Falls back to gt if DT is not
# installed, so the report always renders.
interactive_table <- function(df, caption = NULL, page_length = 15) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(
      df, filter = "top", rownames = FALSE, caption = caption,
      options = list(pageLength = page_length, scrollX = TRUE)
    )
  } else {
    tbl <- gt::gt(df) |> gt::opt_row_striping()
    if (!is.null(caption)) tbl <- tbl |> gt::tab_header(title = caption)
    tbl
  }
}

missing_note <- function(script) {
  cat(paste0(
    "::: {.callout-note}\n## Not yet generated\n",
    "Run `", script, "` and re-render. This section fills itself in.\n:::\n"
  ))
}

T1  <- read_out("bss_b_T1_inventory.csv")
T2  <- read_out("bss_b_T2_variability.csv")
T3  <- read_out("bss_b_T3_pink_effect.csv")
comp <- read_out("bss_b_comparability.csv")
ledger <- read_out("bss_b_run_ledger.csv")
toy_recovery <- read_out("bss_b_toy_recovery.csv")
```

## The question

Freshwater creel estimates from the Bayesian State-Space (BSS) model include an
**effort-index bias term `b`**: `b[1]` scales the vehicle index count, `b[2]`
the trailer index count. A prior of `lognormal(0, sigma)` puts its median at
1. 

`b` is only identified when a fishery-year has **census (tie-in) counts**
alongside its index counts. Without census count data the posterior simply reproduces the
prior, as an assumption, not an estimate.

We expect fishery-years where census data is unavailable, with an immediate resource challenge for 
Snohomish and Stillguamish fall salmon fisheries. So:

> **First-order question: how variable is `b` across years, rivers and
> fisheries?** If it is stable, historical estimates transfer to a year we
> cannot measure. If it is highly variable, they do not, and we need a
> defensible way to construct a predictive distribution instead.

That second case is what §5 and §6 set up.

## How to read this

```{r}
#| results: asis
if (!is.null(ledger)) {
  n_ok <- sum(ledger$status == "ok", na.rm = TRUE)
  n_skip <- sum(ledger$status == "skipped", na.rm = TRUE)
  n_err <- sum(ledger$status == "error", na.rm = TRUE)
  cat(paste0(
    "This run attempted **", nrow(ledger), " fishery-years**: ",
    n_ok, " fitted, ", n_skip, " skipped, ", n_err, " errored. ",
    "Skips are usually structural (no vehicle counts, one angler type) rather than failures -- ",
    "see the run ledger in the appendix for the reason on each.\n\n"
  ))
}
```

Every number here is reproducible from this repository with **no VPN and no
database access** (`DATA_SOURCE <- "external"` reads data.wa.gov; the one
internal-only parameter, the estimation window, is captured in
`lookup/fishery_params.csv`). Clone the branch and re-run `00` → `06`.

## Headline

```{r}
#| results: asis
if (is.null(T2) || is.null(T3)) {
  missing_note("analysis/bss_bias/06_variability_analysis.R")
} else {
  veh2 <- T2 |> filter(bias_type == "vehicle", !is.na(tau))
  veh3 <- T3 |> filter(bias_type == "vehicle")

  if (nrow(veh2) > 0) {
    tau_rng <- range(veh2$tau, na.rm = TRUE)
    # tau is an SD on the log scale, so exp(tau) reads as a multiplicative
    # year-to-year swing -- the most legible form for a mixed audience.
    cat(paste0(
      "- **Interannual variability (b₁, vehicle):** τ ranges ",
      round(tau_rng[1], 3), " to ", round(tau_rng[2], 3),
      " across fishery series (log scale). As a multiplicative swing that is roughly ",
      "×", round(exp(tau_rng[1]), 2), " to ×", round(exp(tau_rng[2]), 2),
      " year to year, beyond what each year's own uncertainty explains.\n"
    ))
    widest <- veh2 |> slice_max(tau, n = 1)
    cat(paste0(
      "- **Widest series:** ", widest$fishery_type[1], " (τ = ", round(widest$tau[1], 3),
      "), prediction interval for a new year ", round(widest$pi_lb[1], 2), " – ",
      round(widest$pi_ub[1], 2), ".\n"
    ))
  }
  if (nrow(veh3) > 0) {
    cat(paste0(
      "- **Pink-year effect (b₁):** ×", round(veh3$pink_ratio[1], 3),
      " (95% CI ", round(veh3$pink_ci_lo[1], 3), " – ", round(veh3$pink_ci_hi[1], 3),
      ", p = ", signif(veh3$pink_p[1], 3), "), explaining ",
      round(veh3$pct_year_var_explained[1], 1), "% of the between-year variance.\n"
    ))
    cat(paste0(
      "- **Where the variance lives (b₁):** between rivers ",
      round(veh3$var_basin[1], 4), ", between fisheries ", round(veh3$var_fishery[1], 4),
      ", between years ", round(veh3$var_year[1], 4), ", estimation error ",
      round(veh3$var_measurement[1], 4), " (log-scale variances).\n"
    ))
  }
}
```

## 1. Inventory of bias terms

Every `b` estimate in hand, by likelihood type, river and year.

::: {.callout-warning}
## Read `prior_contraction` before quoting a number
Prior contraction near 0 means the posterior barely moved off its prior -- the
estimate reflects the assumption, not the data. This matters more for
`b[2]` than `b[1]`. A fishery-year with no trailer index counts still produces a `b[2]`
posterior, but low variance across such years is the prior talking, not a stable
bias term.
:::

```{r}
#| results: asis
if (is.null(T1)) missing_note("analysis/bss_bias/06_variability_analysis.R")
```

```{r}
if (!is.null(T1)) {
  T1 |>
    transmute(
      Basin = basin, Fishery = fishery_type, Year = year_start,
      `Year type` = parity, Likelihood = likelihood,
      b = round(b_median, 3), `q10` = round(q10, 3), `q90` = round(q90, 3),
      `Prior contraction` = round(prior_contraction, 3),
      Flag = informed_flag, Tier = comparability_tier,
      `Window (days)` = n_days_in_window
    ) |>
    interactive_table(page_length = 20)
}
```

## 2. Interannual variability

`b[1]` (vehicle) and `b[2]` (trailer) across years, by basin. Marker shape
encodes the comparability tier; see §7 for what the tiers mean.

::: {.panel-tabset}

### b₁ (vehicle)

```{r}
#| results: asis
p <- fig_path("fig1_b_vehicle_by_year")
if (is.null(p)) missing_note("analysis/bss_bias/03_plot_b_series.R") else cat(paste0("![](", p, ")\n"))
```

### b₂ (trailer)

```{r}
#| results: asis
p <- fig_path("fig1b_b_trailer_by_year")
if (is.null(p)) missing_note("analysis/bss_bias/03_plot_b_series.R") else cat(paste0("![](", p, ")\n"))
```

### Posterior densities

```{r}
#| results: asis
for (nm in c("fig3_posterior_densities", "fig3b_posterior_densities_trailer")) {
  p <- fig_path(nm)
  if (!is.null(p)) cat(paste0("![](", p, ")\n\n"))
}
if (is.null(fig_path("fig3_posterior_densities"))) missing_note("analysis/bss_bias/03_plot_b_series.R")
```

:::

### How much of that is real?

The spread of the point estimates mixes two things: genuine year-to-year
movement in `b`, and uncertainty in measuring each year. **τ² separates them**, 
it is the variance remaining after each estimate's own uncertainty is
accounted for. That is why this uses a random-effects fit rather than the
standard deviation of the point estimates, which would overstate the real
variability.

```{r}
#| results: asis
if (is.null(T2)) missing_note("analysis/bss_bias/06_variability_analysis.R")
```

```{r}
if (!is.null(T2)) {
  T2 |>
    transmute(
      Basin = basin, Fishery = fishery_type, Likelihood = bias_type,
      `N yrs` = n_years, `N informed` = n_informed,
      b = round(pooled_b, 3),
      `CI lo` = round(ci_lb, 3), `CI hi` = round(ci_ub, 3),
      `PI lo` = round(pi_lb, 3), `PI hi` = round(pi_ub, 3),
      `tau` = round(tau, 3), `I2 (%)` = round(I2, 1), `Q p` = signif(Q_p, 3),
      `tau2 (informed only)` = round(tau2_informed_only, 4),
      Note = note
    ) |>
    interactive_table()
}
```

::: {.callout-note}
## CI vs PI
**CI** is the interval on the *mean of past years*. **PI** is where a *new*
year is expected to fall, and is wider by construction. The PI is the one that
answers "what do we use for a year we cannot measure", as quoting the CI in its
place would understate uncertainty.

`tau2 (informed only)` recomputes τ² over data-informed estimates alone. If it
is much larger than `tau`², the apparent stability came from prior-dominated
years sitting near `b = 1`.
:::

## 3. Pink-year effect

Puget Sound pink salmon return in **odd** years, bringing large effort influxes
that plausibly shift the vehicle:angler relationship.

```{r}
#| results: asis
p <- fig_path("fig7_pink_year_effect")
if (is.null(p)) missing_note("analysis/bss_bias/06_variability_analysis.R") else cat(paste0("![](", p, ")\n"))
p2 <- fig_path("fig7b_pink_year_effect_trailer")
if (!is.null(p2)) cat(paste0("\n![](", p2, ")\n"))
```

```{r}
if (!is.null(T3)) {
  T3 |>
    transmute(
      Likelihood = bias_type, `N est` = n_estimates, `N series` = n_series,
      `Between rivers` = round(var_basin, 4),
      `Between fisheries` = round(var_fishery, 4),
      `Between years` = round(var_year, 4),
      `Estimation error` = round(var_measurement, 4),
      `Pink ×b` = round(pink_ratio, 3),
      `CI lo` = round(pink_ci_lo, 3), `CI hi` = round(pink_ci_hi, 3),
      `p` = signif(pink_p, 3),
      `% year var. explained` = round(pct_year_var_explained, 1)
    ) |>
    interactive_table(page_length = 10)
}
```

::: {.callout-warning}
## How much can this contrast actually carry?
The pink effect is estimated from a handful of odd years against a handful of
even ones. Treat a wide CI as what it is, where the design has limited power, not
evidence of no effect. It is a hypothesis worth carrying into the predictive
model, not a settled coefficient.
:::

## 4. What this implies for prediction

```{r}
#| results: asis
p <- fig_path("fig8_predictive_distribution")
if (is.null(p)) missing_note("analysis/bss_bias/06_variability_analysis.R") else cat(paste0("![](", p, ")\n"))
p2 <- fig_path("fig8b_predictive_distribution_trailer")
if (!is.null(p2)) cat(paste0("\n![](", p2, ")\n"))
```

The shaded band is the **95% prediction interval for a year we did not
observe**, built from that fishery's own history. It is the direct answer to
"what do we use when there is no census data, and how uncertain is it?"

- **Tight band** → historical `b` transfers; use the pooled estimate with its PI.
- **Wide band** → it does not transfer, and a single historical value would be
  false precision.

## 5. The conceptual model

A ladder, each rung nesting inside the next. The group's decision is which rung
to stand on.

| Rung | Model | What it buys |
|---|---|---|
| **0** | Use last year's `b` | Status quo. No pooling, no uncertainty propagation. |
| **1** | `log b̂ᵢᵧ ~ N(log θᵢᵧ, seᵢᵧ²)`, `log θᵢᵧ ~ N(μᵢ, τ²)` | τ² is the variability answer; the prediction interval **is** the predictive distribution |
| **2** | `log θᵢᵧ ~ N(μᵢ + β·pinkᵧ, τ²)` | β = pink effect; drop in τ² = how much pink explains |
| **3** | `μᵢ ~ N(M, σ²_river)` | Pools across rivers; a data-poor fishery borrows strength |
| **4** | Hierarchical `b` **inside** the Stan model | Propagates uncertainty exactly instead of in two stages |

::: {.callout-important}
## Rungs 1–3 are an approximation
They take each year's `se` as fixed and known, rather than propagating the full
posterior. That is standard practice and defensible, but it is an
approximation to rung 4, and worth stating plainly rather than discovering
later. Rung 4 is a model rewrite, not a config change.
:::

## 6. Toy example: does the approach actually work?

Before applying this to real numbers, here it is on a **simulated world where
the truth is known** -- three rivers, seven years, a known pink effect and
known between-river and between-year variances.

```{r}
#| results: asis
for (nm in c("figA_toy_simulated_world", "figC_toy_variance_components", "figB_toy_prediction_check")) {
  p <- fig_path(nm)
  if (!is.null(p)) cat(paste0("![](", p, ")\n\n"))
}
if (is.null(fig_path("figA_toy_simulated_world"))) missing_note("analysis/bss_bias/05_predictive_toy.R")
```

```{r}
if (!is.null(toy_recovery)) {
  toy_recovery |>
    gt() |>
    tab_header(title = "Parameter recovery on simulated data",
               subtitle = "Known truth vs. what the method recovered") |>
    opt_row_striping()
}
```

The held-out check is the use case in miniature: drop the most recent year,
refit on the rest, and ask whether the predictive distribution covers the value
that was hidden. **If it does not work here, where the truth is known and the
model is correctly specified, it will not work on the real data.**

## 7. Comparability caveats

Before comparing `b` across years, the group has to agree which fishery-years
are comparable at all.

```{r}
#| results: asis
p <- fig_path("fig2_comparability_strip")
if (is.null(p)) missing_note("analysis/bss_bias/03_plot_b_series.R") else cat(paste0("![](", p, ")\n"))
```

### Season duration and timing

Estimation windows are **not stable across years**. Skagit fall salmon has run
91–140 days; Snohomish fall salmon 33–91; Stillaguamish 2023-24 was 12 days,
its season cut short on Chinook impacts. Start dates move by weeks.

This matters directly for the first-order question: **when the window moves,
some of the year-to-year variation in `b` is a window artifact rather than a
change in vehicle-index bias.** τ² computed across such years mixes the two.

```{r}
#| results: asis
# Render only the columns this file actually has. A bss_b_comparability.csv
# written by an earlier version of 02 lacks the newer duration/timing columns,
# and that is a normal state mid-transition -- it should degrade to a smaller
# table plus a note saying what to re-run, not kill the render.
comp_cols <- c(
  Basin = "basin", Fishery = "fishery_type", Season = "season_label",
  Tier = "comparability_tier", Start = "date_start", End = "date_end",
  Days = "n_days_in_window", `Open days` = "n_days_open",
  `Start vs ref` = "start_shift_days", `Length vs ref` = "pct_of_ref_days",
  `CRC area(s)` = "crc_areas", `Section(s)` = "section_nums",
  `Count types` = "count_types_present"
)

if (!is.null(comp)) {
  absent <- comp_cols[!comp_cols %in% names(comp)]
  if (length(absent) > 0) {
    cat(paste0(
      "::: {.callout-note}\n## Comparability table predates the current pipeline\n",
      "`bss_b_comparability.csv` is missing: `", paste(absent, collapse = "`, `"), "`. ",
      "Re-run `analysis/bss_bias/02_build_comparability_table.R` and re-render to get the full table.\n:::\n\n"
    ))
  }
}
```

```{r}
if (!is.null(comp)) {
  present <- comp_cols[comp_cols %in% names(comp)]
  d <- comp |> select(all_of(present))
  if ("Length vs ref" %in% names(d)) {
    d <- d |> mutate(`Length vs ref` = if_else(is.na(`Length vs ref`), NA_character_,
                                                paste0(round(100 * `Length vs ref`), "%")))
  }
  interactive_table(d, page_length = 20)
}
```

### What the tiers mean

Within each series the **most recent year is the reference**; every other year
is compared against it.

| Tier | Trigger |
|---|---|
| `not-estimable` | No vehicle counts -- `b[1]` is not estimated from data at all |
| `reference` | The anchor year itself |
| `not-comparable` | Study design, count types, or `p_TI` changed vs. reference |
| `comparable-with-caveat` | CRC area / sections / season window shifted, or a thin sample |
| `comparable` | None of the above |

::: {.callout-important}
## These rules are proposals, not findings
Three things worth deciding in the room:

1. **Comparable is the default.** A year is `comparable` unless a specific
   check fires. Anything not on the checklist is invisible to it.
2. **Season duration and timing are reported but do NOT affect the tier.**
   `start_shift_days` and `pct_of_ref_days` are shown above and flagged beyond
   ±14 days / ±25%, but deliberately do not downgrade a year. One line in
   `02_build_comparability_table.R` changes that if the group wants it to.
3. **Variance components in §2–3 are computed over ALL fishery-years**, not
   only comparable ones. A tier-restricted sensitivity run is one filter away
   (`INCLUDE_TIERS` in `06_variability_analysis.R`).
:::

## Appendix

### Diagnostics

```{r}
#| results: asis
for (nm in c("fig5_information_diagnostic", "fig4_vehicle_vs_trailer_bias")) {
  p <- fig_path(nm)
  if (!is.null(p)) cat(paste0("![](", p, ")\n\n"))
}
```

### Run ledger

Status, stage and reason for every fishery-year attempted, with the estimation
window actually used.

```{r}
if (!is.null(ledger)) interactive_table(ledger, page_length = 20)
```

### Reproducing this

```bash
# once, with VPN -- captures the estimation windows, then commit the lookup
Rscript analysis/bss_bias/00b_capture_fishery_params.R

# no VPN needed from here
Rscript analysis/bss_bias/00_discover_fisheries.R
Rscript analysis/bss_bias/01_fit_bss_bias.R          # the slow one (MCMC)
Rscript analysis/bss_bias/02_build_comparability_table.R
Rscript analysis/bss_bias/03_plot_b_series.R
Rscript analysis/bss_bias/05_predictive_toy.R
Rscript analysis/bss_bias/06_variability_analysis.R
quarto render analysis/bss_bias/report/bss_bias_report.qmd
```

### `b` is not `p_TI`in CreelEstimates code 
`b` (this analysis) is the **effort-index bias term**. `p_TI` / `TI_expan` is
the **census:index spatial-coverage ratio** -- a different parameter that sits
adjacent in the same likelihood, and which some code comments confusingly also
call "the bias parameter". `p_TI` appears here only because a year-to-year
change in it would distort `b`; it is not the thing being compared.

