---
title: "BSS effort-bias term (`b`): interannual variability and a predictive approach"
subtitle: "DFW/tribal technical staff working session"
date: last-modified
execute:
  warning: false
  message: false
  echo: false
format:
  html:
    page-layout: full
    code-tools: true
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    title-block-banner: "#1D886E"
---

```{r setup}
library(tidyverse)
library(here)
library(gt)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")
FIG_DIR <- file.path(OUT_DIR, "figures")

# This report is a PRESENTATION LAYER over analysis/bss_bias/outputs/ -- it
# reads CSVs and embeds figures that the numbered scripts already produced,
# and never recomputes anything. That is deliberate: it means the report
# cannot quietly disagree with the pipeline that generated it, it needs no
# database or VPN, and it re-renders in seconds after each refit.

read_out <- function(name) {
  p <- file.path(OUT_DIR, name)
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
}

fig_path <- function(name) {
  p <- file.path(FIG_DIR, paste0(name, ".png"))
  if (file.exists(p)) p else NULL
}

# Interactive where it earns its keep: tables are for interrogating (sort by
# b, filter to one river, search a fishery), so they get DT. Figures are for
# reading, so they are embedded as produced. Falls back to gt if DT is not
# installed, so the report always renders.
interactive_table <- function(df, caption = NULL, page_length = 15) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(
      df, filter = "top", rownames = FALSE, caption = caption,
      options = list(pageLength = page_length, scrollX = TRUE)
    )
  } else {
    tbl <- gt::gt(df) |> gt::opt_row_striping()
    if (!is.null(caption)) tbl <- tbl |> gt::tab_header(title = caption)
    tbl
  }
}

missing_note <- function(script) {
  cat(paste0(
    "::: {.callout-note}\n## Not yet generated\n",
    "Run `", script, "` and re-render. This section fills itself in.\n:::\n"
  ))
}

T1  <- read_out("bss_b_T1_inventory.csv")
T2  <- read_out("bss_b_T2_variability.csv")
T3  <- read_out("bss_b_T3_pink_effect.csv")
comp <- read_out("bss_b_comparability.csv")
ledger <- read_out("bss_b_run_ledger.csv")
```

## The question

Freshwater creel estimates from the Bayesian State-Space (BSS) model include an
**effort-index bias term `b`**: `b[1]` scales the vehicle index count, `b[2]`
the trailer index count. A prior of `lognormal(0, sigma)` puts its median at
1. 

`b` is only identified when a fishery-year has **census (tie-in) counts**
alongside its index counts. Without census count data the posterior simply reproduces the
prior, as an assumption, not an estimate.

We expect fishery-years where census data is unavailable, with an immediate resource challenge for 
Snohomish and Stillaguamish fall salmon fisheries. So:

> **First-order question: how variable is `b` across years, rivers and
> fisheries?** If it is stable, historical estimates transfer to a year we
> cannot measure. If it is highly variable, they do not, and we need a
> defensible way to construct a predictive distribution instead.

That second case is what §4 sets up. The worked predictive distributions and a
known-truth demonstration are in the companion draft,
[proposed next steps](draft/bss_bias_next_steps.html), pending discussion.

### What `b` actually is

The model predicts how many vehicles the index count *should* see, from two
things it estimates separately:

- **angler effort** — how many anglers were present in that section at that
  count, scaled by how much of the section the count covers (`p_TI`)
- **vehicles per angler** (`R_V`) — estimated from the interviews, where each
  interviewed group reports how many anglers it had (`A_A`) and how many
  vehicles it brought (`V_A`)

`b` is the multiplier between that prediction and what the count actually
recorded:

$$
b_1 \;=\; \frac{\text{vehicles actually counted}}{\text{anglers present} \;\times\; \text{section coverage} \;\times\; \text{vehicles per angler}}
$$

So `b = 1` means the index count is exactly what angler effort and the
interviews say it should be. `b[2]` is the same quantity for trailer counts,
using trailers per angler (`R_T`).

| `b` | What it says | Examples |
|---|---|---|
| **above 1** | The count sees **more** vehicles than the anglers account for | Non-angler vehicles inside the counted area: day-use visitors, hikers, swimmers, non-fishing boaters, shared trailhead or boat-launch lots. Vehicles belonging to anglers who fished outside the count window but whose car is still parked. Overnight or multi-day vehicles. The same vehicle counted in two sections or on two passes. |
| **below 1** | The count sees **fewer** vehicles than the anglers account for | Anglers parking outside the count route: roadside pullouts, private drives, neighborhoods, walk-in access. Anglers dropped off, or arriving by boat from a launch outside the section. Parking areas the route misses. Counts timed away from peak vehicle presence. |

Note that `b` is **not** "anglers per vehicle" — that is `R_V`, and it is
estimated separately. `b` is only the leftover discrepancy after `R_V` has
been accounted for.

::: {.callout-note}
## Why `b` needs interviews to exist at all
`b` and `R_V` both multiply angler effort in the same count likelihood, so from
the counts alone they cannot be told apart — any count is equally well explained
by "more vehicles per angler" or "a biased count". The interviews are what pin
`R_V` down independently (`V_A ~ Binomial(A_A, R_V)`), which is what leaves `b`
free to absorb the remaining discrepancy.

A fishery-year with few vehicle-count interviews therefore has little
separating the two, and `b` falls back toward its prior. That is exactly what
the `prior_contraction` column reports.
:::

## How to read this

```{r}
#| results: asis
if (!is.null(ledger)) {
  n_ok <- sum(ledger$status == "ok", na.rm = TRUE)
  n_skip <- sum(ledger$status == "skipped", na.rm = TRUE)
  n_err <- sum(ledger$status == "error", na.rm = TRUE)
  cat(paste0(
    "This run attempted **", nrow(ledger), " fishery-years**: ",
    n_ok, " fitted, ", n_skip, " skipped, ", n_err, " errored. ",
    "Skips are usually structural (no vehicle counts, one angler type) rather than failures -- ",
    "see the run ledger in the appendix for the reason on each.\n\n"
  ))
}
```

Every number here is reproducible from this repository with **no VPN and no
database access** (`DATA_SOURCE <- "external"` reads data.wa.gov; the one
internal-only parameter, the estimation window, is captured in
`lookup/fishery_params.csv`). Clone the branch and re-run `00` → `06`.

## Headline

```{r}
#| results: asis
if (is.null(T2) || is.null(T3)) {
  missing_note("analysis/bss_bias/06_variability_analysis.R")
} else {
  veh2 <- T2 |> filter(bias_type == "vehicle", !is.na(tau))
  veh3 <- T3 |> filter(bias_type == "vehicle")

  if (nrow(veh2) > 0) {
    tau_rng <- range(veh2$tau, na.rm = TRUE)
    # tau is an SD on the log scale, so exp(tau) reads as a multiplicative
    # year-to-year swing -- the most legible form for a mixed audience.
    cat(paste0(
      "- **Interannual variability (b₁, vehicle):** τ ranges ",
      round(tau_rng[1], 3), " to ", round(tau_rng[2], 3),
      " across fishery series (log scale). As a multiplicative swing that is roughly ",
      "×", round(exp(tau_rng[1]), 2), " to ×", round(exp(tau_rng[2]), 2),
      " year to year, beyond what each year's own uncertainty explains.\n"
    ))
    widest <- veh2 |> slice_max(tau, n = 1)
    cat(paste0(
      "- **Widest series:** ", widest$fishery_type[1], " (τ = ", round(widest$tau[1], 3),
      "), prediction interval for a new year ", round(widest$pi_lb[1], 2), " – ",
      round(widest$pi_ub[1], 2), ".\n"
    ))
  }
  if (nrow(veh3) > 0) {
    cat(paste0(
      "- **Pink-year effect (b₁):** ×", round(veh3$pink_ratio[1], 3),
      " (95% CI ", round(veh3$pink_ci_lo[1], 3), " – ", round(veh3$pink_ci_hi[1], 3),
      ", p = ", signif(veh3$pink_p[1], 3), "), explaining ",
      round(veh3$pct_year_var_explained[1], 1), "% of the between-year variance.\n"
    ))
    cat(paste0(
      "- **Where the variance lives (b₁):** between rivers ",
      round(veh3$var_basin[1], 4), ", between fisheries ", round(veh3$var_fishery[1], 4),
      ", between years ", round(veh3$var_year[1], 4), ", estimation error ",
      round(veh3$var_measurement[1], 4), " (log-scale variances).\n"
    ))
  }
}
```

## 1. Inventory of bias terms

Every `b` estimate in hand, by likelihood type, river and year.

::: {.callout-warning}
## Read `prior_contraction` before quoting a number
Prior contraction near 0 means the posterior barely moved off its prior -- the
estimate reflects the assumption, not the data. This matters more for
`b[2]` than `b[1]`. A fishery-year with no trailer index counts still produces a `b[2]`
posterior, but low variance across such years is the prior talking, not a stable
bias term.
:::

```{r}
#| results: asis
if (is.null(T1)) missing_note("analysis/bss_bias/06_variability_analysis.R")
```

```{r}
if (!is.null(T1)) {
  T1 |>
    transmute(
      Basin = basin, Fishery = fishery_type, Year = year_start,
      `Year type` = parity, Likelihood = likelihood,
      b = round(b_median, 3), `q10` = round(q10, 3), `q90` = round(q90, 3),
      `Prior contraction` = round(prior_contraction, 3),
      Flag = informed_flag,
      `Window (days)` = n_days_in_window
    ) |>
    interactive_table(page_length = 20)
}
```

## 2. Interannual variability

`b[1]` (vehicle) and `b[2]` (trailer) across years, by basin. Marker shape
encodes the comparability tier; see §5 for what the tiers mean.

::: {.panel-tabset}

### b₁ (vehicle)

```{r}
#| results: asis
p <- fig_path("fig1_b_vehicle_by_year")
if (is.null(p)) missing_note("analysis/bss_bias/03_plot_b_series.R") else cat(paste0("![](", p, ")\n"))
```

### b₂ (trailer)

```{r}
#| results: asis
p <- fig_path("fig1b_b_trailer_by_year")
if (is.null(p)) missing_note("analysis/bss_bias/03_plot_b_series.R") else cat(paste0("![](", p, ")\n"))
```

### Posterior densities

```{r}
#| results: asis
for (nm in c("fig3_posterior_densities", "fig3b_posterior_densities_trailer")) {
  p <- fig_path(nm)
  if (!is.null(p)) cat(paste0("![](", p, ")\n\n"))
}
if (is.null(fig_path("fig3_posterior_densities"))) missing_note("analysis/bss_bias/03_plot_b_series.R")
```

:::

### How much of that is real?

The spread of the point estimates mixes two things: genuine year-to-year
movement in `b`, and uncertainty in measuring each year. **τ² separates them**, 
it is the variance remaining after each estimate's own uncertainty is
accounted for. That is why this uses a random-effects fit rather than the
standard deviation of the point estimates, which would overstate the real
variability.

```{r}
#| results: asis
if (is.null(T2)) missing_note("analysis/bss_bias/06_variability_analysis.R")
```

```{r}
if (!is.null(T2)) {
  T2 |>
    transmute(
      Basin = basin, Fishery = fishery_type, Likelihood = bias_type,
      `Yrs with an estimate` = n_years_available, `Yrs used` = n_years_used,
      `N informed` = n_informed,
      b = round(pooled_b, 3),
      `CI lo` = round(ci_lb, 3), `CI hi` = round(ci_ub, 3),
      `PI lo` = round(pi_lb, 3), `PI hi` = round(pi_ub, 3),
      `tau` = round(tau, 3), `I2 (%)` = round(I2, 1), `Q p` = signif(Q_p, 3),
      `tau2 (informed only)` = round(tau2_informed_only, 4),
      Note = note
    ) |>
    interactive_table()
}
```

::: {.callout-note}
## CI vs PI
**CI** is the interval on the *mean of past years*. **PI** is where a *new*
year is expected to fall, and is wider by construction. The PI is the one that
answers "what do we use for a year we cannot measure", as quoting the CI in its
place would understate uncertainty.

`tau2 (informed only)` recomputes τ² over data-informed estimates alone. If it
is much larger than `tau`², the apparent stability came from prior-dominated
years sitting near `b = 1`.
:::

## 3. Pink-year effect?

Puget Sound pink salmon return in **odd** years, bringing large effort influxes
that plausibly shift the vehicle:angler relationship.

**This section uses only the fall-timed fisheries.** A fishery has to be
fishing while the pinks are in the river for a pink year to move its effort, so
only Skagit fall salmon, Snohomish fall salmon and Stillaguamish salmon and
gamefish can carry the test. Skagit spring Chinook (Apr–Jul) and summer sockeye
(Jun–Jul) have no in-season overlap with the pink return and are excluded —
including them would not test the hypothesis, only dilute it with years whose
odd/even split is noise. §1 and §2 still cover every fishery.

```{r}
#| results: asis
p <- fig_path("fig7_pink_year_effect")
if (is.null(p)) missing_note("analysis/bss_bias/06_variability_analysis.R") else cat(paste0("![](", p, ")\n"))
p2 <- fig_path("fig7b_pink_year_effect_trailer")
if (!is.null(p2)) cat(paste0("\n![](", p2, ")\n"))
```

```{r}
if (!is.null(T3)) {
  T3 |>
    transmute(
      Likelihood = bias_type, `N est` = n_estimates, `N series` = n_series,
      `Between rivers` = round(var_basin, 4),
      `Between fisheries` = round(var_fishery, 4),
      `Between years` = round(var_year, 4),
      `Estimation error` = round(var_measurement, 4),
      `Pink ×b` = round(pink_ratio, 3),
      `CI lo` = round(pink_ci_lo, 3), `CI hi` = round(pink_ci_hi, 3),
      `p` = signif(pink_p, 3),
      `% year var. explained` = round(pct_year_var_explained, 1)
    ) |>
    interactive_table(page_length = 10)
}
```

::: {.callout-warning}
## How much can this contrast actually carry?
The pink effect is estimated from a handful of odd years against a handful of
even ones. Treat a wide CI as what it is, where the design has limited power, not
evidence of no effect. It is a hypothesis worth carrying into the predictive
model, not a settled coefficient.
:::

## 4. The conceptual model

A ladder, each rung nesting inside the next. The group's decision is which rung
to stand on.

| Rung | In plain terms | Model | What it buys |
|---|---|---|---|
| **0** | `b this year = b last year` | Use last year's `b` | Status quo. No pooling, no uncertainty propagation. |
| **1** | `b ~ fishery + year-to-year scatter` | `log b̂ᵢᵧ ~ N(log θᵢᵧ, seᵢᵧ²)`, `log θᵢᵧ ~ N(μᵢ, τ²)` | τ² is the variability answer; the prediction interval **is** the predictive distribution |
| **2** | `b ~ fishery + pink year + year-to-year scatter` | `log θᵢᵧ ~ N(μᵢ + β·pinkᵧ, τ²)` | β = pink effect; drop in τ² = how much pink explains |
| **3** | `b ~ river + fishery + pink year + year-to-year scatter` | `μᵢ ~ N(M, σ²_river)` | Pools across rivers; a data-poor fishery borrows strength |
| **4** | the same, but estimated *with* the creel model rather than after it | Hierarchical `b` **inside** the Stan model | Propagates uncertainty exactly instead of in two stages |

::: {.callout-important}
## Rungs 1–3 are an approximation
They take each year's `se` as fixed and known, rather than propagating the full
posterior. That is standard practice and defensible, but it is an
approximation to rung 4, and worth stating plainly rather than discovering
later. Rung 4 is a model rewrite, not a config change.
:::

## 5. Is it the same fishery each year?

Before comparing `b` across years, the group has to agree the years are
comparable at all. Two things decide that, and they are not the same:

- **what the program defined** — the water bodies, sections and survey sites in
  the location lookup;
- **what the survey record actually holds** — the days worked, and among those,
  the days carrying *both* an index count and the census count it pairs with.

The last of those is the one that governs `b`. `b` is a single pooled scalar
(`vector[G] b` in the Stan model, with no section or day index), so it is
informed **only** by section-days holding both counts — not by every surveyed
day, and not at all by the size of the defined footprint.

```{r}
#| results: asis
p <- fig_path("fig16_fishery_composition")
if (is.null(p)) missing_note("analysis/bss_bias/02b_survey_coverage.R") else cat(paste0("![](", p, ")\n"))
```

Read the rows top to bottom: sections defined → sections surveyed → days
surveyed → paired days. **A gap opening between consecutive rows is where a
plan stopped becoming data.**

```{r}
composition <- read_out("bss_b_fishery_composition.csv")
if (!is.null(composition)) {
  composition |>
    transmute(
      Fishery = fishery_type, Year = year, Water = water_bodies,
      `Sections defined` = sections_defined, `Index sites` = index_sites_defined,
      `Window days` = n_days_window, `Sections surveyed` = sections_surveyed,
      `Days surveyed` = n_days_surveyed, `Paired index+census days` = n_days_tie_in
    ) |>
    interactive_table(page_length = 20)
} else {
  missing_note("analysis/bss_bias/02b_survey_coverage.R")
}
```

### What this does to the variability question

Two consequences worth carrying into every number above.

**The evidence behind `b` is thinner than the survey record suggests.** Skagit
fall salmon surveyed 50 days in 2024 and produced 7 paired ones; Stillaguamish
surveyed 33 days in 2022-23 for 2. The markers in Figures 1 and 1b are sized by
this count, so a point resting on almost nothing looks like it.

**Estimation windows are not stable across years.** Skagit fall salmon has run
91–140 days, Snohomish fall salmon 33–91, and Stillaguamish 2023-24 was 12 days
with its season cut short on Chinook impacts. Because `b` is pooled across all
days, a moved window changes which period it averages — so some year-to-year
movement in `b` is a window artifact rather than a change in vehicle-index bias,
and τ² computed across such years mixes the two.

That second point bears directly on §3: in Skagit fall salmon and Snohomish
fall salmon, **every odd year starts in mid-August and every even year on 1
September**, so a pink-year coefficient in those basins would be estimating
"August plus pink". `bss_b_parity_confound.csv` states this per fishery, and
`bss_b_common_window_cost.csv` gives the window all years share along with what
refitting there would cost in paired days.


::: {.callout-important}
## What still needs deciding in the room
1. **Nothing here downgrades a year automatically.** The window and coverage
   differences above are reported, not scored. Whether a year with 2 paired
   days belongs in the same τ² as one with 18 is a judgement the group has to
   make, and it is not made anywhere in this pipeline.
2. **Variance components in §2–3 are computed over ALL fishery-years.** A
   restricted sensitivity run is one filter away (`INCLUDE_TIERS` in
   `06_variability_analysis.R`), and `bss_b_fishery_composition.csv` is the
   table to pick that restriction from.
3. **Section numbering is not stable between years.** Blocks get split and
   everything above them renumbered, so a section number is a label rather
   than a place. `bss_b_lut_block_lineage.csv` matches blocks between years on
   their site membership and separates a renumbering from a genuine redraw.
:::

## Appendix

### Diagnostics

```{r}
#| results: asis
for (nm in c("fig5_information_diagnostic", "fig4_vehicle_vs_trailer_bias")) {
  p <- fig_path(nm)
  if (!is.null(p)) cat(paste0("![](", p, ")\n\n"))
}
```

### Run ledger

Status, stage and reason for every fishery-year attempted, with the estimation
window actually used.

```{r}
if (!is.null(ledger)) interactive_table(ledger, page_length = 20)
```

### Reproducing this

```bash
# once, with VPN -- captures the estimation windows, then commit the lookup
Rscript analysis/bss_bias/00b_capture_fishery_params.R

# no VPN needed from here
Rscript analysis/bss_bias/00_discover_fisheries.R
Rscript analysis/bss_bias/01_fit_bss_bias.R          # the slow one (MCMC)
Rscript analysis/bss_bias/02_build_comparability_table.R
Rscript analysis/bss_bias/03_plot_b_series.R
Rscript analysis/bss_bias/06_variability_analysis.R
quarto render analysis/bss_bias/report/bss_bias_report.qmd
```

The companion draft (predictive distributions and the known-truth
demonstration) additionally needs `05_predictive_toy.R`; see
`report/draft/bss_bias_next_steps.qmd`.

### `b` is not `p_TI` in CreelEstimates code
`b` (this analysis) is the **effort-index bias term**. `p_TI` / `TI_expan` is
the **census:index spatial-coverage ratio** -- a different parameter that sits
adjacent in the same likelihood, and which some code comments confusingly also
call "the bias parameter". `p_TI` appears here only because a year-to-year
change in it would distort `b`; it is not the thing being compared.

### How multi-season fisheries are classified odd vs. even

Pink-year parity keys off the **fall run year**, not the calendar span in the
name. `Stillaguamish salmon and gamefish 2022-23` is a fall-2022 fishery, so it
counts as **even** (non-pink); `2023-24` counts as odd. Concretely, parity is
taken from `year_start`, the first year in the name.

This decides which side of the pink contrast the Stillaguamish multi-season
years land on, so it is worth being explicit about rather than leaving to
inference from the figures.

