---
title: "Using prior census data to set the effort-index bias term"
date: 2026-09-16
format:
  docx:
    # Calibri, black, one body size. Built by make_reference_docx.R -- run that
    # once and commit the result; without the file the docx render fails.
    reference-doc: reference-plain.docx
  html:
    embed-resources: true
    toc: false
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
# ==============================================================================
# The memo's numbers come from the modelling outputs, not from the prose.
#
# Nothing below is typed in. The b estimates and intervals are read from
# bss_b_summary.csv (written by 01_fit_bss_bias.R) and the survey counts from
# bss_b_scope_effort.csv (12_fork_scope_compare.R). Re-render after a refit
# and the memo is current.
#
# kable(), not gt(): gt does not render to Word, and this memo is edited in
# Word. The same file produces .docx and .html.
# ==============================================================================
library(tidyverse)
library(knitr)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")

read_if <- function(f) {
  p <- file.path(OUT_DIR, f)
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
}

b_all  <- read_if("bss_b_summary.csv")
effort <- read_if("bss_b_scope_effort.csv")

FITS <- tibble::tribble(
  ~fit,              ~scope, ~fishery_name,
  "Mainstem 2024",   "MS",   "Stillaguamish salmon and gamefish 2024-25",
  "Mainstem 2025",   "MS",   "Stillaguamish salmon and gamefish 2025-26",
  "North Fork 2024", "NF",   "Stillaguamish salmon and gamefish 2024-25",
  "North Fork 2025", "NF",   "Stillaguamish salmon and gamefish 2025-26"
) |>
  mutate(
    out_name = paste0(fishery_name, " [", scope, "]"),
    # 12_fork_scope_compare.R keys its outputs on "<scope> <year>" -- "MS 2024"
    # -- not on the display label used in this memo. Joining the counts on
    # `fit` therefore matched nothing and every count column rendered as an em
    # dash while the file sat there full of numbers. Derived the same way 12
    # derives it, so the two cannot drift.
    scope_fit = paste(scope, as.integer(str_extract(fishery_name, "\\d{4}")))
  )

# A fallback of tibble(fit = character()) joins cleanly but carries none of
# the columns the chunks then reference, so a missing outputs file failed with
# "object 'n_paired_anchors' not found" instead of rendering em dashes -- the
# exact case the fallback exists to cover. These carry the columns, empty.
empty_like <- function(key, num_cols) {
  out <- tibble::tibble(.rows = 0)
  out[[key]] <- character()
  for (n in num_cols) out[[n]] <- numeric()
  out
}

effort_or_empty <- function() {
  effort %||% empty_like("fit", c("n_index_sites", "n_index_days", "n_paired_anchors",
                                  "census_anglers_bank", "census_anglers_boat"))
}

b_or_empty <- function() {
  if (!is.null(b_all)) return(b_all)
  empty_like("fishery_name", c("median", "q2.5", "q97.5")) |>
    dplyr::mutate(bias_type = character())
}

fmt <- function(x, d = 2) ifelse(is.na(x), "—", formatC(x, format = "f", digits = d))
n_fmt <- function(x) ifelse(is.na(x), "—", formatC(x, format = "d", big.mark = ","))

have_fits <- !is.null(b_all) &&
  any(FITS$out_name %in% b_all$fishery_name[b_all$bias_type == "vehicle"])
```

Technical memo describing proposed modifications to effort
bias-correction approach for Snohomish and Stillaguamish basin
freshwater fishery monitoring for 2026.

Following the elimination of WDFW's District 13 freshwater salmon
harvest monitoring budget, aerial census effort counts (also known as tie-in
counts) are not funded for Stillaguamish mainstem salmon and
North Fork Stillaguamish gamefish fisheries in 2026. These surveys are
used to estimate a census-index bias parameter (b), which corrects for
under or overcounts of anglers obtained during index counts of vehicles
and boat trailers at access sites. Measuring it requires census counts
run alongside index counts on the same day, time, and river section.
Without either in-season census counts or an informative prior on the
creel model's b parameter, the index counts of vehicles and trailers
would be assumed to be an unbiased measure of fishing effort. Where
census are reduced or eliminated this season, WDFW proposes to use
historical data, where available and appropriate, for census-index effort
bias terms from prior years to inform the bias correction used in the current year.

## Proposed action

We propose to generate bias terms for vehicle and trailer index counts from the fishery’s own prior years. The distribution of a historical bias term should be wide enough to span variability observed in the past. Following a review of b parameter estimates encompassing entire fishery seasons, we evaluated the same parameter on a subset of data for years 2024 and 2025 in the mainstem and North Fork Stillaguamish Rivers to achieve a comparable temporal and spatial footprint of the 2026 fishery.

The two most recent Stillaguamish seasons were compared on a common window, September 16th to October 31st, restricted to sections in the mainstem and North Fork Stillaguamish Rivers. The same reaches were surveyed in both years, from the same access sites and on the same number of paired census days (@tbl-survey).


```{r pending-note, results = "asis"}
if (!have_fits) {
  cat("TO BE COMPLETED when the current fits finish. Bias term values below are",
      "not yet available.\n")
}
```



```{r}
#| label: tbl-survey
#| tbl-cap: "Survey effort behind each bias term, on the common September 16 to October 31 window."
counts <- FITS |>
  left_join(
    effort_or_empty() |>
      select(fit, n_index_sites, n_index_days, n_paired_anchors,
             census_anglers_bank, census_anglers_boat),
    by = c("scope_fit" = "fit")
  ) |>
  transmute(
    Fit = fit,
    `Index sites`  = n_fmt(n_index_sites),
    `Index days`   = n_fmt(n_index_days),
    `Census days`  = n_fmt(n_paired_anchors),
    `Bank anglers` = n_fmt(census_anglers_bank),
    `Boat anglers` = n_fmt(census_anglers_boat)
  )

kable(counts)
```

{{< pagebreak >}}

Estimated bias terms for both index types are given in @tbl-b.

```{r}
#| label: tbl-b
#| tbl-cap: "Estimated census-index bias terms with 95 percent credible intervals."
# BOTH index types, one row each, rather than a vehicle-only table with the
# trailer term suppressed underneath it. A wide interval is itself the finding
# and reads more honestly than an absence, so the table reports the trailer
# term and lets the interval carry the caveat.
#
# Long rather than wide: eight rows of four narrow columns, not four rows of
# six. Word wraps a numeric header one character per line when the table runs
# past the text block.
b_tbl <- FITS |>
  tidyr::crossing(bias_type = c("vehicle", "trailer")) |>
  left_join(
    b_or_empty() |> select(out_name = fishery_name, bias_type, median, q2.5, q97.5),
    by = c("out_name", "bias_type")
  ) |>
  arrange(match(fit, FITS$fit), desc(bias_type)) |>
  transmute(
    Fit = fit,
    Index = if_else(bias_type == "vehicle", "Vehicle", "Trailer"),
    b = fmt(median),
    `95 percent interval` = ifelse(is.na(median), "—",
                                   paste0(fmt(q2.5), "–", fmt(q97.5)))
  )

kable(b_tbl)
```

@fig-b shows the same estimates on a log scale, where the width of an
interval is comparable between terms of different size.

&nbsp;

```{r}
#| label: fig-b
#| fig-cap: "Bias term estimates and 95 percent credible intervals. The dashed line marks b = 1, where index counts are an unbiased measure of effort."
#| fig-width: 6.5
#| fig-height: 2.8
#| fig-dpi: 300
# fig-dpi stays inside the `#|` block: Quarto reads cell options only up to the
# first line that is not `#|`, so a plain comment in the middle of them drops
# everything below it. Word embeds a raster and the default is 96 dpi, which is
# what made the figure grainy on the page.
#
# Built here from bss_b_summary.csv rather than read as an image, so the figure
# refreshes with the fits and the memo stays self-contained.
#
# Log x: b is multiplicative -- catch scales as 1/b -- so a factor-of-two
# difference should look the same whether it sits at 0.5 or at 4. On a linear
# axis the trailer intervals would flatten the vehicle ones to nothing.
fig_dat <- FITS |>
  tidyr::crossing(bias_type = c("vehicle", "trailer")) |>
  left_join(
    b_or_empty() |> select(out_name = fishery_name, bias_type, median, q2.5, q97.5),
    by = c("out_name", "bias_type")
  ) |>
  filter(!is.na(median)) |>
  mutate(
    Index = factor(if_else(bias_type == "vehicle", "Vehicle", "Trailer"),
                   levels = c("Vehicle", "Trailer")),
    fit   = factor(fit, levels = rev(FITS$fit))
  )

if (nrow(fig_dat) > 0) {
  ggplot(fig_dat, aes(x = median, y = fit)) +
    geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
    geom_errorbarh(aes(xmin = q2.5, xmax = q97.5), height = 0.18) +
    geom_point(size = 2) +
    facet_wrap(~Index) +
    scale_x_log10() +
    labs(x = "Bias term b", y = NULL) +
    theme_bw(base_size = 10) +
    # theme_bw() sets axis text to grey30. Every piece of type in the figure is
    # black, to match the document.
    theme(panel.grid.minor = element_blank(),
          text        = element_text(colour = "black"),
          axis.text   = element_text(colour = "black"),
          axis.title  = element_text(colour = "black"),
          strip.text  = element_text(colour = "black"),
          legend.text = element_text(colour = "black"),
          strip.background = element_rect(fill = "white", colour = "black"))
}
```

For 2026 vehicle and trailer bias terms in the creel model, we propose placing a lognormal prior on b, using the mean and variance of the b parameters for vehicle and trailer counts from that fishery’s own prior years (@tbl-b, @fig-b). Adjusting code and model structure may be needed to accommodate this modification. This action requires an additional assumption that the index to census relationship is stable enough between the prior years and 2026 for a past b to be representative, and that a b parameter based on historical data sufficiently captures the parameter’s inter annual variability. If any census data is collected this season, the prior and the census data would combine, and the census will update whenever there is adequate data. This work would proceed in collaboration with Stillaguamish and Snohomish Tribal technical staff, with all code and supporting documentation made available for all participants. 
