---
title: "Using prior census data to set the effort-index bias term"
date: 2026-09-16
format:
  docx: default
  html:
    embed-resources: true
    toc: false
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
# ==============================================================================
# The memo's numbers come from the modelling outputs, not from the prose.
#
# Nothing below is typed in. The b estimates and intervals are read from
# bss_b_summary.csv (written by 01_fit_bss_bias.R), the survey counts from
# bss_b_scope_effort.csv and bss_b_scope_sites.csv (12_fork_scope_compare.R),
# and the census reach names and river miles from the committed location
# lookup. Re-render after a refit and the memo is current.
#
# kable(), not gt(): gt does not render to Word, and this memo is edited in
# Word. The same file produces .docx and .html.
# ==============================================================================
library(tidyverse)
library(knitr)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")
LUT     <- here::here("analysis", "bss_bias", "lookup", "fishery_location_lut.csv")

# Boat anglers in census below which the trailer term is called data-limited.
# Same value 12_fork_scope_compare.R uses, restated rather than sourced so the
# memo renders without running that script.
CENSUS_BOAT_FLOOR <- 20

read_if <- function(f) {
  p <- file.path(OUT_DIR, f)
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
}

b_all  <- read_if("bss_b_summary.csv")
effort <- read_if("bss_b_scope_effort.csv")
sites  <- read_if("bss_b_scope_sites.csv")

FITS <- tibble::tribble(
  ~fit,              ~scope, ~fishery_name,                                ~water_body,
  "Mainstem 2024",   "MS",   "Stillaguamish salmon and gamefish 2024-25",  "Stillaguamish - MS",
  "Mainstem 2025",   "MS",   "Stillaguamish salmon and gamefish 2025-26",  "Stillaguamish - MS",
  "North Fork 2024", "NF",   "Stillaguamish salmon and gamefish 2024-25",  "Stillaguamish - NF",
  "North Fork 2025", "NF",   "Stillaguamish salmon and gamefish 2025-26",  "Stillaguamish - NF"
) |>
  mutate(out_name = paste0(fishery_name, " [", scope, "]"))

# Census reach names and river miles, from the committed lookup. Names rather
# than section numbers because the numbers are not stable between years: the
# same mainstem water is sections 2 and 3 in 2024 and 1 and 2 in 2025.
reaches <- if (file.exists(LUT)) {
  readr::read_csv(LUT, show_col_types = FALSE) |>
    filter(location_type == "Section") |>
    distinct(fishery_name, water_body_code, section_num, location_code, lower_rm, upper_rm) |>
    mutate(across(c(lower_rm, upper_rm), ~suppressWarnings(as.numeric(.x))))
} else NULL

reach_label <- function(fn, wb) {
  if (is.null(reaches)) return(NA_character_)
  r <- reaches |> filter(fishery_name == fn, water_body_code == wb) |> arrange(lower_rm)
  if (nrow(r) == 0) return(NA_character_)
  paste(sprintf("%s (%.3g–%.3g)", r$location_code, r$lower_rm, r$upper_rm), collapse = "; ")
}

fmt <- function(x, d = 2) ifelse(is.na(x), "—", formatC(x, format = "f", digits = d))
n_fmt <- function(x) ifelse(is.na(x), "—", formatC(x, format = "d", big.mark = ","))

have_fits <- !is.null(b_all) &&
  any(FITS$out_name %in% b_all$fishery_name[b_all$bias_type == "vehicle"])
```

Technical memo describing proposed modifications to effort
bias-correction approach for Snohomish and Stillaguamish basin
freshwater fishery monitoring for 2026.

Following the elimination of WDFW's District 13 freshwater salmon
harvest monitoring budget, aerial census counts (also known as tie-in
counts) are not funded for the Snohomish and Stillaguamish mainstem and
North Fork Stillaguamish gamefish fishery in 2026. These surveys are
used to estimate a census-index bias parameter (b), which corrects for
under- or overcounts of anglers obtained during index counts of vehicles
and boat trailers at access sites. Measuring it requires census counts
run alongside index counts on the same day, time, and river section.
Without either in-season census counts or an informative prior on the
creel model's b parameter, the index counts of vehicles and trailers
would be assumed to be an unbiased measure of fishing effort. Where
census are reduced or eliminated this season, WDFW proposes to use
historical data for census-index effort bias terms from prior years to
inform the bias correction used in the current year.

## Proposed action

Use bias terms for vehicle and trailer index counts from the fishery's
own prior years, matching a comparable temporal and spatial footprint to
the 2026 fishery, with a distribution wide enough to span the
year-to-year variation observed.

```{r pending-note, results = "asis"}
if (!have_fits) {
  cat("TO BE COMPLETED when the current fits finish. Bias term values below are",
      "not yet available.\n")
}
```

The two most recent Stillaguamish seasons were compared on a common
window, September 16th to October 31st, restricted to sections in the
mainstem and North Fork Stillaguamish Rivers .

```{r survey-table}
survey <- FITS |>
  mutate(`Census reach (river mile)` = map2_chr(fishery_name, water_body, reach_label)) |>
  left_join(
    if (is.null(effort)) tibble(fit = character()) else
      effort |> select(fit, n_index_sites, n_index_days, n_paired_anchors,
                       census_anglers_bank, census_anglers_boat),
    by = "fit"
  ) |>
  transmute(
    Fit = fit,
    `Census reach (river mile)`,
    `Index sites`          = n_fmt(n_index_sites),
    `Index days`           = n_fmt(n_index_days),
    `Paired census days`   = n_fmt(n_paired_anchors),
    `Census anglers, bank` = n_fmt(census_anglers_bank),
    `Census anglers, boat` = n_fmt(census_anglers_boat)
  )

kable(survey)
```

The index sites are the same named access points in both years. The
upper North Fork above Cicero Bridge and the South Fork were surveyed in
2024 and not in 2025, and are excluded from both years for that reason.

```{r b-table}
b_tbl <- FITS |>
  left_join(
    if (is.null(b_all)) tibble(out_name = character()) else
      b_all |> filter(bias_type == "vehicle") |>
        select(out_name = fishery_name, median, q2.5, q97.5, informed_flag,
               prior_contraction),
    by = "out_name"
  ) |>
  left_join(
    if (is.null(effort)) tibble(fit = character()) else
      effort |> select(fit, census_anglers_boat, n_paired_anchors),
    by = "fit"
  ) |>
  mutate(
    limitation = case_when(
      is.na(median)                      ~ "Not yet fitted",
      informed_flag == "unconverged"     ~ "Did not converge",
      informed_flag == "prior-dominated" ~ "Posterior tracks the prior",
      informed_flag == "weak"            ~ "Few angler interviews",
      TRUE                               ~ paste0(n_paired_anchors, " paired census days")
    )
  ) |>
  transmute(
    Fit = fit,
    `Vehicle b` = fmt(median),
    `95 percent interval` = ifelse(is.na(median), "—",
                                   paste0(fmt(q2.5), "–", fmt(q97.5))),
    `Data limitation` = limitation
  )

kable(b_tbl)
```

```{r trailer-note, results = "asis"}
if (!is.null(effort)) {
  thin <- effort |> filter(census_anglers_boat < CENSUS_BOAT_FLOOR)
  if (nrow(thin) > 0) {
    cat(sprintf(
      paste0("Trailer bias terms are not reported. Fewer than %d boat anglers were ",
             "counted in census in each of these fishery-years (%s), and the trailer ",
             "index is measured against boat anglers in census rather than against ",
             "trailers counted.\n"),
      CENSUS_BOAT_FLOOR,
      paste(sprintf("%s: %d", thin$fit, as.integer(thin$census_anglers_boat)), collapse = "; ")
    ))
  }
}
```

## Proposed approach for this season

**Vehicle term.** Place a lognormal prior on b, centred on the pooled
estimate from that fishery's own prior years and given a spread taken
from the observed year-to-year variation in those same estimates,
widened by the leave-one-out prediction interval for that series where
more than two years are available.

The distinction matters: a single year's posterior can be narrow while
the year-to-year variation is wide, and it is the year-to-year variation
that describes what an unmeasured year might be.

```{r spread-evidence, results = "asis"}
ms <- FITS |>
  filter(scope == "MS") |>
  left_join(
    if (is.null(b_all)) tibble(out_name = character()) else
      b_all |> filter(bias_type == "vehicle") |>
        select(out_name = fishery_name, median, q2.5, q97.5, informed_flag),
    by = "out_name"
  ) |>
  filter(!is.na(median), informed_flag != "unconverged")

if (nrow(ms) == 2) {
  lo <- ms |> slice_min(median); hi <- ms |> slice_max(median)
  cat(sprintf(
    paste0("The mainstem is the clearest case. Its vehicle term was %s in %s and %s in ",
           "%s, on the same two reaches, the same %s access sites, and three paired ",
           "census days in each year. The two 95 percent intervals (%s–%s and ",
           "%s–%s) do not overlap. Carrying the %s value into %s unchanged would ",
           "have understated estimated catch and effort by about %.0f percent, because ",
           "catch scales as 1/b.\n"),
    fmt(hi$median), str_extract(hi$fit, "\\d{4}"),
    fmt(lo$median), str_extract(lo$fit, "\\d{4}"),
    if (is.null(effort)) "same" else
      n_fmt(effort$n_index_sites[match(hi$fit, effort$fit)]),
    fmt(hi$q2.5), fmt(hi$q97.5), fmt(lo$q2.5), fmt(lo$q97.5),
    str_extract(hi$fit, "\\d{4}"), str_extract(lo$fit, "\\d{4}"),
    100 * (1 - lo$median / hi$median)
  ))
}
```

**Trailer term.** Fix at 1 where census boat-angler coverage falls below
the threshold used in the comparison above. This is a stated assumption,
not an estimate, and it is reported as such. In a fishery that is 2 to 3
percent boat effort, the choice moves the season total very little, and
the alternative is a number with no data behind it.

**Where census is collected this season.** The prior and the census data
combine in the usual way, and the census will dominate wherever it is
adequate. The prior is what keeps the estimate stable where it is not.

```{r sources, results = "asis"}
stamp <- file.info(file.path(OUT_DIR, "bss_b_summary.csv"))$mtime
cat("\n---\n\n*Bias term estimates and survey counts in this memo are read directly from ",
    "the modelling outputs in `analysis/bss_bias/outputs/`",
    if (!is.na(stamp)) paste0(" (fits last written ", format(stamp, "%Y-%m-%d %H:%M"), ")") else "",
    ". Re-render to refresh.*\n", sep = "")
```
