---
title: "Using prior census data to set the effort-index bias term"
date: 2026-09-16
format:
  docx:
    # Calibri, black, one body size. Built by make_reference_docx.R -- run that
    # once and commit the result; without the file the docx render fails.
    reference-doc: reference-plain.docx
  html:
    embed-resources: true
    toc: false
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
# ==============================================================================
# The memo's numbers come from the modelling outputs, not from the prose.
#
# Nothing below is typed in. The b estimates and intervals are read from
# bss_b_summary.csv (written by 01_fit_bss_bias.R), the survey counts from
# bss_b_scope_effort.csv and bss_b_scope_sites.csv (12_fork_scope_compare.R),
# and the census reach names and river miles from the committed location
# lookup. Re-render after a refit and the memo is current.
#
# kable(), not gt(): gt does not render to Word, and this memo is edited in
# Word. The same file produces .docx and .html.
# ==============================================================================
library(tidyverse)
library(knitr)

OUT_DIR <- here::here("analysis", "bss_bias", "outputs")
LUT     <- here::here("analysis", "bss_bias", "lookup", "fishery_location_lut.csv")

# Boat anglers in census below which the trailer term is called data-limited.
# Same value 12_fork_scope_compare.R uses, restated rather than sourced so the
# memo renders without running that script.
CENSUS_BOAT_FLOOR <- 20

read_if <- function(f) {
  p <- file.path(OUT_DIR, f)
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
}

b_all  <- read_if("bss_b_summary.csv")
effort <- read_if("bss_b_scope_effort.csv")
sites  <- read_if("bss_b_scope_sites.csv")

FITS <- tibble::tribble(
  ~fit,              ~scope, ~fishery_name,                                ~water_body,
  "Mainstem 2024",   "MS",   "Stillaguamish salmon and gamefish 2024-25",  "Stillaguamish - MS",
  "Mainstem 2025",   "MS",   "Stillaguamish salmon and gamefish 2025-26",  "Stillaguamish - MS",
  "North Fork 2024", "NF",   "Stillaguamish salmon and gamefish 2024-25",  "Stillaguamish - NF",
  "North Fork 2025", "NF",   "Stillaguamish salmon and gamefish 2025-26",  "Stillaguamish - NF"
) |>
  mutate(
    out_name = paste0(fishery_name, " [", scope, "]"),
    # 12_fork_scope_compare.R keys its outputs on "<scope> <year>" -- "MS 2024"
    # -- not on the display label used in this memo. Joining the counts on
    # `fit` therefore matched nothing and every count column rendered as an em
    # dash while the file sat there full of numbers. Derived the same way 12
    # derives it, so the two cannot drift.
    scope_fit = paste(scope, as.integer(str_extract(fishery_name, "\\d{4}")))
  )

# Census reach names and river miles, from the committed lookup. Names rather
# than section numbers because the numbers are not stable between years: the
# same mainstem water is sections 2 and 3 in 2024 and 1 and 2 in 2025.
#
# RESTRICTED TO THE SECTIONS ACTUALLY FITTED. Listing every section of a fork
# put four census blocks and 37 river miles against North Fork 2024, which is
# held to section 4 alone -- the table then described water the estimate does
# not cover. scope_rules.R is the same file the fits read, so the two cannot
# disagree.
source(here::here("analysis", "bss_bias", "scope_rules.R"))

reaches <- if (file.exists(LUT)) {
  readr::read_csv(LUT, show_col_types = FALSE) |>
    filter(location_type == "Section") |>
    distinct(fishery_name, water_body_code, section_num, location_code, lower_rm, upper_rm) |>
    mutate(section_num = suppressWarnings(as.numeric(section_num)),
           across(c(lower_rm, upper_rm), ~suppressWarnings(as.numeric(.x))))
} else NULL

fitted_sections <- function(fn, tag) {
  # scope_rules.R narrates what it drops through cli, which knitr would lift
  # into the rendered table. Capture it: the rule is what we want, not its
  # commentary.
  RUN_SCOPE <<- SCOPE_PRESETS[[tag]]
  out <- NULL
  invisible(utils::capture.output(
    out <- tryCatch(suppressMessages(fishery_section_limit(fn)),
                    error = function(e) NULL),
    type = "output"))
  RUN_SCOPE <<- NULL
  out
}

reach_rows <- function(fit, fn, wb, tag) {
  if (is.null(reaches)) return(NULL)
  keep <- fitted_sections(fn, tag)
  r <- reaches |> filter(fishery_name == fn, water_body_code == wb)
  if (!is.null(keep)) r <- r |> filter(section_num %in% keep)
  if (nrow(r) == 0) return(NULL)
  r |> arrange(lower_rm) |>
    transmute(Fit = fit,
              `Census reach` = location_code,
              `River mile` = paste0(signif(lower_rm, 3), "-", signif(upper_rm, 3)))
}

fmt <- function(x, d = 2) ifelse(is.na(x), "—", formatC(x, format = "f", digits = d))
n_fmt <- function(x) ifelse(is.na(x), "—", formatC(x, format = "d", big.mark = ","))

have_fits <- !is.null(b_all) &&
  any(FITS$out_name %in% b_all$fishery_name[b_all$bias_type == "vehicle"])
```

Technical memo describing proposed modifications to effort
bias-correction approach for Snohomish and Stillaguamish basin
freshwater fishery monitoring for 2026.

Following the elimination of WDFW's District 13 freshwater salmon
harvest monitoring budget, aerial census effort counts (also known as tie-in
counts) are not funded for the Snohomish and Stillaguamish mainstem and
North Fork Stillaguamish gamefish fishery in 2026. These surveys are
used to estimate a census-index bias parameter (b), which corrects for
under or overcounts of anglers obtained during index counts of vehicles
and boat trailers at access sites. Measuring it requires census counts
run alongside index counts on the same day, time, and river section.
Without either in-season census counts or an informative prior on the
creel model's b parameter, the index counts of vehicles and trailers
would be assumed to be an unbiased measure of fishing effort. Where
census are reduced or eliminated this season, WDFW proposes to use
historical data for census-index effort bias terms from prior years to
inform the bias correction used in the current year.

## Proposed action

We propose to use bias terms for vehicle and trailer index counts from the fishery's
own prior years, matched to a comparable temporal and spatial footprint of
the 2026 fishery. The distribution of a historical bias term should be wide enough to span the
year-to-year variation observed in the past.

```{r pending-note, results = "asis"}
if (!have_fits) {
  cat("TO BE COMPLETED when the current fits finish. Bias term values below are",
      "not yet available.\n")
}
```

The two most recent Stillaguamish seasons were compared on a common
window, September 16th to October 31st, restricted to sections in the
mainstem and North Fork Stillaguamish Rivers. The same reaches were surveyed in both years. 

```{r reach-table}
# TWO NARROW TABLES, not one wide one. A single table carrying the reach names
# alongside six count columns forced Word to wrap every numeric header one
# character per line.
reach_tbl <- bind_rows(pmap(
  list(FITS$fit, FITS$fishery_name, FITS$water_body, FITS$scope), reach_rows))

if (nrow(reach_tbl) > 0) kable(reach_tbl)
```

```{r count-table}
counts <- FITS |>
  left_join(
    if (is.null(effort)) tibble(fit = character()) else
      effort |> select(fit, n_index_sites, n_index_days, n_paired_anchors,
                       census_anglers_bank, census_anglers_boat),
    by = c("scope_fit" = "fit")
  ) |>
  transmute(
    Fit = fit,
    `Index sites`  = n_fmt(n_index_sites),
    `Index days`   = n_fmt(n_index_days),
    `Census days`  = n_fmt(n_paired_anchors),
    `Bank anglers` = n_fmt(census_anglers_bank),
    `Boat anglers` = n_fmt(census_anglers_boat)
  )

kable(counts)
```

```{r b-table}
b_tbl <- FITS |>
  left_join(
    if (is.null(b_all)) tibble(out_name = character()) else
      b_all |> filter(bias_type == "vehicle") |>
        select(out_name = fishery_name, median, q2.5, q97.5, informed_flag,
               prior_contraction),
    by = "out_name"
  ) |>
  left_join(
    if (is.null(effort)) tibble(fit = character()) else
      effort |> select(fit, census_anglers_boat, n_paired_anchors),
    by = c("scope_fit" = "fit")
  ) |>
  mutate(
    limitation = case_when(
      is.na(median)                      ~ "Not yet fitted",
      informed_flag == "unconverged"     ~ "Did not converge",
      informed_flag == "prior-dominated" ~ "Posterior tracks the prior",
      informed_flag == "weak"            ~ "Few angler interviews",
      is.na(n_paired_anchors)            ~ "not yet summarised",
      TRUE                               ~ paste0(n_paired_anchors, " paired census days")
    )
  ) |>
  transmute(
    Fit = fit,
    `Vehicle b` = fmt(median),
    `95 percent interval` = ifelse(is.na(median), "—",
                                   paste0(fmt(q2.5), "–", fmt(q97.5))),
    `Data limitation` = limitation
  )

kable(b_tbl)
```

```{r trailer-note, results = "asis"}
if (!is.null(effort)) {
  # Through FITS, so the prose names the fishery-years the same way the tables
  # above do rather than falling back to the "MS 2024" scope tag.
  thin <- FITS |>
    inner_join(effort |> select(fit, census_anglers_boat),
               by = c("scope_fit" = "fit")) |>
    filter(census_anglers_boat < CENSUS_BOAT_FLOOR)
  if (nrow(thin) > 0) {
    cat(sprintf(
      paste0("Trailer bias terms are not reported. Fewer than %d boat anglers were ",
             "counted in census in each of these fishery-years (%s), and the trailer ",
             "index is measured against boat anglers in census rather than against ",
             "trailers counted.\n"),
      CENSUS_BOAT_FLOOR,
      paste(sprintf("%s: %d", thin$fit, as.integer(thin$census_anglers_boat)), collapse = "; ")
    ))
  }
}
```

## Proposed approach for this season

For 2026 vehicle and trailer bias terms in the creel model, place a lognormal prior on b, centered on the pooled
estimate from that fishery's own prior years and given a spread taken
from the observed year-to-year variation in those same estimates. 

```{r spread-evidence, results = "asis"}
ms <- FITS |>
  filter(scope == "MS") |>
  left_join(
    if (is.null(b_all)) tibble(out_name = character()) else
      b_all |> filter(bias_type == "vehicle") |>
        select(out_name = fishery_name, median, q2.5, q97.5, informed_flag),
    by = "out_name"
  ) |>
  filter(!is.na(median), informed_flag != "unconverged")

if (nrow(ms) == 2) {
  lo <- ms |> slice_min(median); hi <- ms |> slice_max(median)
  cat(sprintf(
    paste0("The mainstem is the clearest case. Its vehicle term was %s in %s and %s in ",
           "%s, on the same two reaches, the same %s access sites, and three paired ",
           "census days in each year. The two 95 percent intervals (%s–%s and ",
           "%s–%s) do not overlap. Carrying the %s value into %s unchanged would ",
           "have understated estimated catch and effort by about %.0f percent, because ",
           "catch scales as 1/b.\n"),
    fmt(hi$median), str_extract(hi$fit, "\\d{4}"),
    fmt(lo$median), str_extract(lo$fit, "\\d{4}"),
    if (is.null(effort)) "same" else
      n_fmt(effort$n_index_sites[match(hi$fit, effort$fit)]),
    fmt(hi$q2.5), fmt(hi$q97.5), fmt(lo$q2.5), fmt(lo$q97.5),
    str_extract(hi$fit, "\\d{4}"), str_extract(lo$fit, "\\d{4}"),
    100 * (1 - lo$median / hi$median)
  ))
}
```

If any census data is collected this season, the prior and the census data
would combine, and the census will update whenever there is
adequate data.

```{r sources, results = "asis"}
stamp <- file.info(file.path(OUT_DIR, "bss_b_summary.csv"))$mtime
cat("\n---\n\n*Bias term estimates and survey counts in this memo are read directly from ",
    "the modelling outputs in `analysis/bss_bias/outputs/`",
    if (!is.na(stamp)) paste0(" (fits last written ", format(stamp, "%Y-%m-%d %H:%M"), ")") else "",
    ". Re-render to refresh.*\n", sep = "")
```
